---
title: "THEM6: GSEA analysis"
author: "Peter Repiscak"
#date: '`r format(Sys.time(), "%d %B %Y")`'
output:
 html_document:
  code_folding: hide
  toc: true
  toc_float: false
  theme: united
always_allow_html: true
---

# Brief description
Gene Set Enrichment Analysis of Them6 differential expression results from 4 conditions (AI_WT, AI_CTL, AI_KO1, AI_KO2) and following comparisons:

- KO1_vs_CTL
- KO2_vs_CTL
- KO1_vs_WT
- KO2_vs_WT

GSEA was perfomed in discovery mode (using gs_hallmark, gs_C2_kegg, gs_C2_reactome, gs_C5_GO(BP|CC|MF)) and supervised mode to assess lipid, sterol or cholesterol pathway down-regulation in THEM6_KO.


# Methods
For gene set enrichment analysis (GSEA), genes from differential expression analysis were ranked according to log2FC and GSEA was carried out on ranked list using clusterProfiler v`r packageVersion("clusterProfiler")` with fgsea v`r packageVersion("fgsea")` algorithm. The GSEA enrichment was done against the GOBP STEROL HOMEOSTASIS gene set obtained from the Molecular Signatures Database (MSigDB) (Liberzon *et al*, 2011) using msigdbr v`r packageVersion("msigdbr")` package. Final GSEA plots were generated using modified version of gseaplot2 from enrichplot package.

```{r analysis_parameters, include=FALSE}
# Specifying parameters here ----
if(!("import" %in% installed.packages())){install.packages("import")}
if(!("renv" %in% installed.packages())){remotes::install_github("rstudio/renv")}

# rewrite below!
base_dir <- "/home/rstudio/workspace/"
data_dir <- file.path(base_dir, "datasets") 
results_dir <- file.path(base_dir, "results_dir") 
dir.create(results_dir)
setwd(base_dir)

#project setup
project_name <- "Them6"
nthreads <- 4  # for furrr multisession

# this make take a while...
# https://rstudio.github.io/renv/reference/config.html
# getOption(x, default = NULL)
#renv::settings$use.cache()
# getOption('renv.config.pak.enabled')
# if the project does not automatically activate run:
if(Sys.getenv("RENV_PATHS_CACHE") != "/renv_cache") {Sys.setenv(RENV_PATHS_CACHE = "/renv_cache")}
if(Sys.getenv("RENV_PATHS_LIBRARY") != "/home/rstudio/renv_library") {Sys.setenv(RENV_PATHS_LIBRARY = "/home/rstudio/renv_library")}
#if(Sys.getenv("RENV_CONFIG_PAK_ENABLED") != "TRUE") {Sys.setenv(RENV_CONFIG_PAK_ENABLED = TRUE)} # ad pak, targets and benchmarkme to docker!!!
# setting root dir
# 0. renv::activate
renv::activate(project = base_dir)
# DO NOT RUN:
renv:init()
# FOR the first run restore environment and packages from them6_gsea_renv.lock 
# 1. restore original environment
renv::restore(project = base_dir, lockfile = file.path(base_dir, "them6_gsea_renv.lock"), prompt = FALSE)
#file.copy(from = file.path(base_dir, "them6_gsea_renv.lock"), to = file.path(base_dir, "renv.lock")) # to use with renv::diagnostics()
```

```{r initial_setup, include=FALSE}
set.seed(42)

import::from(knitr, opts_chunk)

# report options ----
# knitr::opts_chunk
opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      cache.lazy = FALSE,
                      message=FALSE,
                      warning=FALSE,
                      dev = "png")

# cache.path = file.path(results_dir,"report","cache/")
# fig.path = file.path(results_dir,"report","files/")                     
                     
options(width=100)

```

```{r run in the terminal, eval = FALSE, include = FALSE}
# run in terminal to generate report
rmarkdown::render(output_file = file.path(base_dir, "them6_gsea.html"), 
                  input = here::here("them6_gsea.Rmd"))

# rmarkdown::render(input = "/home/peter_r/analysis/them6_gsea.Rmd", output_file = stringr::str_replace_all(date(),c(" +"="-", ":"="_")), output_dir = file.path("/home/peter_r/analysis/report","html/"))
```

```{r libraries, message=FALSE}
# importing only key functions that are actually used - not to polute namespace!
import::from(readr, read_csv)
import::from(magrittr, "%>%")
import::from(dplyr, mutate, select, filter, rename, arrange, desc, group_by, summarise, ungroup)  # dplyr_mutate = mutate
import::from(purrr, map)
import::from(future, plan, multisession, sequential)
import::from(furrr, furrr_options, future_map2)
import::from(ggplot2, .all=TRUE) # importing all as there is too many
#import::from(.from = DOSE, gseaScores)
             
# importing functions ----
#https://import.rticulate.org/articles/import.html#advanced-usage-1
# renv::install('bit64')
import::from(.from = here::here("scripts/utils.R"), "save_rds", "digest_match", "create_rnk_list", "gsea_function", "gseaPlotESprofile", "tableGrob2", "gsInfo", .character_only=TRUE)
import:::from(gseaScores, .from = DOSE)

# if there is an issue with clusterProfiler: "object 'get_fun_from_pkg' not found"
# downgrade "rvcheck":
# renv::install("rvcheck@0.1.8") rather than devtools::install_version("rvcheck", version = "0.1.8", repos = "http://cran.us.r-project.org")
# and re-install renv::install("bioc::clusterProfiler") rather than BiocManager::install("clusterProfiler")

```

```{r loading datasets and genesets, include=FALSE, eval=TRUE}
# loading datasets ----
# md5_digest them6_deseq2_selColumns_datasets.rds: 806547a11fb22309f4515f31033d6fed
raw_datasets <- readRDS(file.path(data_dir, "them6_deseq2_selColumns_datasets.rds"))
# digest::digest(raw_datasets, algo = "md5")
message("Correct data file loaded: ", digest_match(x = raw_datasets, digestfile = file.path(data_dir, "them6_deseq2_selColumns_datasets_digest.md5")))

# check if correct data is loaded in case of missing _digest.md5
if (digest::digest(raw_datasets, algo = "md5") != "806547a11fb22309f4515f31033d6fed") {
  warning("raw_datasets md5 is: ", digest::digest(raw_datasets, algo = "md5"), " and it differs from expected: 806547a11fb22309f4515f31033d6fed")
}

# loading genesets ----
gs_hallmark <- msigdbr::msigdbr(species = "Homo sapiens", category = c("H")) %>%
  dplyr::distinct(gs_name, gene_symbol) %>% as.data.frame()  #dplyr::select(gs_name, entrez_gene)
gs_C2_kegg <- msigdbr::msigdbr(species = "Homo sapiens", category = c("C2"), subcategory = "CP:KEGG") %>%
  dplyr::distinct(gs_name, gene_symbol) %>% as.data.frame()
gs_C2_reactome <- msigdbr::msigdbr(species = "Homo sapiens", category = c("C2"), subcategory = "CP:REACTOME") %>%
  dplyr::distinct(gs_name, gene_symbol) %>% as.data.frame()
gs_C5_GOBP <- msigdbr::msigdbr(species = "Homo sapiens", category = c("C5"), subcategory = "GO:BP") %>%
  dplyr::distinct(gs_name, gene_symbol) %>% as.data.frame()
gs_C5_GOCC <- msigdbr::msigdbr(species = "Homo sapiens", category = c("C5"), subcategory = "GO:CC") %>%
  dplyr::distinct(gs_name, gene_symbol) %>% as.data.frame()
gs_C5_GOMF <- msigdbr::msigdbr(species = "Homo sapiens", category = c("C5"), subcategory = "GO:MF") %>%
  dplyr::distinct(gs_name, gene_symbol) %>% as.data.frame()

gs_symbol_collection <- list(hallmark = gs_hallmark,
                             C2_kegg = gs_C2_kegg,
                             C2_reactome = gs_C2_reactome,
                             C5_GOBP = gs_C5_GOBP,
                             C5_GOCC = gs_C5_GOCC,
                             C5_GOMF = gs_C5_GOMF)

```

```{r Running GSEA on selected genesets, include=FALSE, eval=TRUE}
# generating list of ranked lists (for each of the comparisons)
rnk_lists <- map(names(raw_datasets), ~create_rnk_list(data_name = .x, raw_datasets = raw_datasets)) %>% setNames(names(raw_datasets))

# running GSEA analysis ----
if (!file.exists(file.path(results_dir, "them6_gsea_results.rds"))) {
  message("Generating GSEA results...")
  
  furrr_map_options <- furrr_options(seed = 123)
  if(nthreads > 1) {plan(multisession, workers = nthreads)} else {plan(sequential)} 
  
  KO1_vs_CTL_gsea_results <- furrr::future_map2(.x = gs_symbol_collection,
                                                .y = names(gs_symbol_collection),
                                                .f = function(geneset, geneset_name)
                                                  gsea_function(geneset=geneset,
                                                                geneset_name=geneset_name,
                                                                ranked_list=rnk_lists$KO1_vs_CTL),
                                                .options = furrr_map_options,
                                                .progress = FALSE) 
  
  KO2_vs_CTL_gsea_results <- furrr::future_map2(.x = gs_symbol_collection,
                                                .y = names(gs_symbol_collection),
                                                .f = function(geneset, geneset_name)
                                                  gsea_function(geneset=geneset,
                                                                geneset_name=geneset_name,
                                                                ranked_list=rnk_lists$KO2_vs_CTL),
                                                .options = furrr_map_options,
                                                .progress = FALSE) 
  
  KO1_vs_WT_gsea_results <- furrr::future_map2(.x = gs_symbol_collection,
                                               .y = names(gs_symbol_collection),
                                               .f = function(geneset, geneset_name)
                                                 gsea_function(geneset=geneset,
                                                               geneset_name=geneset_name,
                                                               ranked_list=rnk_lists$KO1_vs_WT),
                                               .options = furrr_map_options,
                                               .progress = FALSE) 
  
  KO2_vs_WT_gsea_results <- furrr::future_map2(.x = gs_symbol_collection,
                                               .y = names(gs_symbol_collection),
                                               .f = function(geneset, geneset_name)
                                                 gsea_function(geneset=geneset,
                                                               geneset_name=geneset_name,
                                                               ranked_list=rnk_lists$KO2_vs_WT),
                                               .options = furrr_map_options,
                                               .progress = FALSE) 
  
  plan(sequential, .cleanup = TRUE)
  
  
  # packing results and saving
  gsea_results <- list(KO1_vs_CTL_gsea_results = KO1_vs_CTL_gsea_results,
                        KO2_vs_CTL_gsea_results = KO2_vs_CTL_gsea_results,
                        KO1_vs_WT_gsea_results = KO1_vs_WT_gsea_results,
                        KO2_vs_WT_gsea_results = KO2_vs_WT_gsea_results)

  save_rds(x = gsea_results,
           rdsname = file.path(results_dir, "them6_gsea_results.rds"), 
           digestname = file.path(results_dir, "them6_gsea_results_digest.md5"))
     

} else {
  message("GSEA results available...loading!")
  # md5_digest them6_gsea_results.rds: 707217f5b76a5db6a36526ac6fce78c9
  gsea_results <- readRDS(file.path(results_dir, "them6_gsea_results.rds")) 
  message("Correct data file loaded: ", digest_match(x = gsea_results, digestfile = file.path(results_dir, "them6_gsea_results_digest.md5")))
  # check if correct data is loaded in case of missing _digest.md5
  if (digest::digest(gsea_results, algo = "md5") != "707217f5b76a5db6a36526ac6fce78c9") {
    warning("gsea_results md5 is: ", digest::digest(gsea_results, algo = "md5"), " and it differs from expected: 707217f5b76a5db6a36526ac6fce78c9")
  }
  
  KO1_vs_CTL_gsea_results <- gsea_results[["KO1_vs_CTL_gsea_results"]]
  KO2_vs_CTL_gsea_results <- gsea_results[["KO2_vs_CTL_gsea_results"]]
  KO1_vs_WT_gsea_results <- gsea_results[["KO1_vs_WT_gsea_results"]]
  KO2_vs_WT_gsea_results <- gsea_results[["KO2_vs_WT_gsea_results"]]

}

# converting gseaResult to data.frames ----
KO1_vs_CTL_gsea_results_df <- map(KO1_vs_CTL_gsea_results, as.data.frame) %>% setNames(names(KO1_vs_CTL_gsea_results))
KO2_vs_CTL_gsea_results_df <- map(KO2_vs_CTL_gsea_results, as.data.frame) %>% setNames(names(KO2_vs_CTL_gsea_results))
KO1_vs_WT_gsea_results_df <- map(KO1_vs_WT_gsea_results, as.data.frame) %>% setNames(names(KO1_vs_WT_gsea_results))
KO2_vs_WT_gsea_results_df <- map(KO2_vs_WT_gsea_results, as.data.frame) %>% setNames(names(KO2_vs_WT_gsea_results))

# saving data.frames as results ----
openxlsx::write.xlsx(KO1_vs_CTL_gsea_results_df, file = file.path(results_dir, "KO1_vs_CTL_gsea.xlsx")) 
openxlsx::write.xlsx(KO2_vs_CTL_gsea_results_df, file = file.path(results_dir, "KO2_vs_CTL_gsea.xlsx")) 
openxlsx::write.xlsx(KO1_vs_WT_gsea_results_df, file = file.path(results_dir, "KO1_vs_WT_gsea.xlsx")) 
openxlsx::write.xlsx(KO2_vs_WT_gsea_results_df, file = file.path(results_dir, "KO2_vs_WT_gsea.xlsx")) 
```

```{r Running GSEA on LIPID/STEROL genesets, include=FALSE, eval=TRUE}
# lipid metabolism ----
# GOBP LIPID and GOBP STEROL
gs_C5_GOBP_LIPID <- gs_C5_GOBP %>%
  dplyr::filter(grepl(pattern = "LIPID", x = gs_name))

gs_C5_GOBP_STEROL <- gs_C5_GOBP %>%
  dplyr::filter(grepl(pattern = "STEROL", x = gs_name))

gs_symbol_collection_LIPID_STEROL <- list(LIPID = gs_C5_GOBP_LIPID,
                             STEROL = gs_C5_GOBP_STEROL)

# running GSEA analysis ----
if (!file.exists(file.path(results_dir, "them6_gsea_lipids_results.rds"))) {
  
  furrr_map_options <- furrr_options(seed = 123)
  if(nthreads > 1) {plan(multisession, workers = nthreads)} else {plan(sequential)} 
  
  KO1_vs_CTL_gsea_LIPIDS_results <- furrr::future_map2(.x = gs_symbol_collection_LIPID_STEROL,
                                                       .y = names(gs_symbol_collection_LIPID_STEROL),
                                                       .f = function(geneset, geneset_name)
                                                         gsea_function(geneset=geneset,
                                                                       geneset_name=geneset_name,
                                                                       ranked_list=rnk_lists$KO1_vs_CTL),
                                                       .options = furrr_map_options,
                                                       .progress = FALSE) 
  
  KO2_vs_CTL_gsea_LIPIDS_results <- furrr::future_map2(.x = gs_symbol_collection_LIPID_STEROL,
                                                       .y = names(gs_symbol_collection_LIPID_STEROL),
                                                       .f = function(geneset, geneset_name)
                                                         gsea_function(geneset=geneset,
                                                                       geneset_name=geneset_name,
                                                                       ranked_list=rnk_lists$KO2_vs_CTL),
                                                       .options = furrr_map_options,
                                                       .progress = FALSE) 
  
  KO1_vs_WT_gsea_LIPIDS_results <- furrr::future_map2(.x = gs_symbol_collection_LIPID_STEROL,
                                                      .y = names(gs_symbol_collection_LIPID_STEROL),
                                                      .f = function(geneset, geneset_name)
                                                        gsea_function(geneset=geneset,
                                                                      geneset_name=geneset_name,
                                                                      ranked_list=rnk_lists$KO1_vs_WT),
                                                      .options = furrr_map_options,
                                                      .progress = FALSE) 
  
  KO2_vs_WT_gsea_LIPIDS_results <- furrr::future_map2(.x = gs_symbol_collection_LIPID_STEROL,
                                                      .y = names(gs_symbol_collection_LIPID_STEROL),
                                                      .f = function(geneset, geneset_name)
                                                        gsea_function(geneset=geneset,
                                                                      geneset_name=geneset_name,
                                                                      ranked_list=rnk_lists$KO2_vs_WT),
                                                      .options = furrr_map_options,
                                                      .progress = FALSE) 
  
  plan(sequential, .cleanup = TRUE)
  
    # packing results and saving
  gsea_lipid_results <- list(KO1_vs_CTL_gsea_LIPIDS_results = KO1_vs_CTL_gsea_LIPIDS_results,
                             KO2_vs_CTL_gsea_LIPIDS_results = KO2_vs_CTL_gsea_LIPIDS_results,
                             KO1_vs_WT_gsea_LIPIDS_results = KO1_vs_WT_gsea_LIPIDS_results,
                             KO2_vs_WT_gsea_LIPIDS_results = KO2_vs_WT_gsea_LIPIDS_results)

  save_rds(x = gsea_lipid_results,
           rdsname = file.path(results_dir, "them6_gsea_lipids_results.rds"), 
           digestname = file.path(results_dir, "them6_gsea_lipids_results_digest.md5"))
     
} else {
  message("GSEA results available...loading!")
  # md5_digest them6_gsea_lipids_results.rds: a71b7c35fddb1442dfbfabbfd77d1105
  gsea_lipid_results <- readRDS(file.path(results_dir, "them6_gsea_lipids_results.rds")) 
  message("Correct data file loaded: ", digest_match(x = gsea_lipid_results, digestfile = file.path(results_dir, "them6_gsea_lipids_results_digest.md5")))
  # check if correct data is loaded in case of missing _digest.md5
  if (digest::digest(gsea_lipid_results, algo = "md5") != "a71b7c35fddb1442dfbfabbfd77d1105") {
    warning("gsea_lipid_results md5 is: ", digest::digest(gsea_lipid_results, algo = "md5"), " and it differs from expected: a71b7c35fddb1442dfbfabbfd77d1105")
  }
  
  KO1_vs_CTL_gsea_LIPIDS_results <- gsea_lipid_results[["KO1_vs_CTL_gsea_LIPIDS_results"]]
  KO2_vs_CTL_gsea_LIPIDS_results <- gsea_lipid_results[["KO2_vs_CTL_gsea_LIPIDS_results"]]
  KO1_vs_WT_gsea_LIPIDS_results <- gsea_lipid_results[["KO1_vs_WT_gsea_LIPIDS_results"]]
  KO2_vs_WT_gsea_LIPIDS_results <- gsea_lipid_results[["KO2_vs_WT_gsea_LIPIDS_results"]]
}

# converting gseaResult to data.frames ----
KO1_vs_CTL_gsea_LIPIDS_results_df <- map(KO1_vs_CTL_gsea_LIPIDS_results, as.data.frame) %>% setNames(names(KO1_vs_CTL_gsea_LIPIDS_results))
KO2_vs_CTL_gsea_LIPIDS_results_df <- map(KO2_vs_CTL_gsea_LIPIDS_results, as.data.frame) %>% setNames(names(KO2_vs_CTL_gsea_LIPIDS_results))
KO1_vs_WT_gsea_LIPIDS_results_df <- map(KO1_vs_WT_gsea_LIPIDS_results, as.data.frame) %>% setNames(names(KO1_vs_WT_gsea_LIPIDS_results))
KO2_vs_WT_gsea_LIPIDS_results_df <- map(KO2_vs_WT_gsea_LIPIDS_results, as.data.frame) %>% setNames(names(KO2_vs_WT_gsea_LIPIDS_results))

# saving data.frames as results ----
openxlsx::write.xlsx(KO1_vs_CTL_gsea_LIPIDS_results_df, file = file.path(results_dir, "KO1_vs_CTL_gsea_LIPIDS.xlsx")) 
openxlsx::write.xlsx(KO2_vs_CTL_gsea_LIPIDS_results_df, file = file.path(results_dir, "KO2_vs_CTL_gsea_LIPIDS.xlsx")) 
openxlsx::write.xlsx(KO1_vs_WT_gsea_LIPIDS_results_df, file = file.path(results_dir, "KO1_vs_WT_gsea_LIPIDS.xlsx")) 
openxlsx::write.xlsx(KO2_vs_WT_gsea_LIPIDS_results_df, file = file.path(results_dir, "KO2_vs_WT_gsea_LIPIDS.xlsx")) 

```

# GSEA results {.tabset}

Results are generated in the results_dir. This folder contains complete results saved as .RData/.xlsx files and figures saved as .tiff. 

```{r GSEA results plots, include=FALSE, eval=TRUE}
# plotting GSEA plots for 

# KO1_vs_CTL_gsea_LIPIDS_results_df
# # LIPID:
# GOBP_REGULATION_OF_LIPID_METABOLIC_PROCESS
# GOBP_REGULATION_OF_LIPID_BIOSYNTHETIC_PROCESS
# GOBP_LIPID_HOMEOSTASIS
# STEROL:
# GOBP_STEROL_HOMEOSTASIS
# GOBP_REGULATION_OF_CHOLESTEROL_METABOLIC_PROCESS
# 
# KO2_vs_CTL_gsea_LIPIDS_results_df
# LIPID:
# GOBP_REGULATION_OF_LIPID_METABOLIC_PROCESS
# GOBP_REGULATION_OF_LIPID_BIOSYNTHETIC_PROCESS
# STEROL:
# GOBP_STEROL_HOMEOSTASIS

# individual plots ----

# genesets of interest
lipid_genesets <- c("GOBP_REGULATION_OF_LIPID_METABOLIC_PROCESS", "GOBP_REGULATION_OF_LIPID_BIOSYNTHETIC_PROCESS", "GOBP_LIPID_HOMEOSTASIS")
KO1_vs_CTL_sterol_genesets <- c("GOBP_STEROL_HOMEOSTASIS", "GOBP_REGULATION_OF_CHOLESTEROL_METABOLIC_PROCESS")
KO2_vs_CTL_sterol_genesets <- c("GOBP_STEROL_HOMEOSTASIS")

# KO1_vs_CTL_gsea_LIPIDS_results
KO1_vs_CTL_LIPID_gseaplots <- purrr::map(.x = lipid_genesets, .f = function(geneset_id){
  gseaPlotESprofile(KO1_vs_CTL_gsea_LIPIDS_results$LIPID, geneSetID = geneset_id, ES_geom="line", NES_table = FALSE, remove_substring = NULL, title = "")
}) %>% setNames(lipid_genesets)

KO1_vs_CTL_STEROL_gseaplots <- purrr::map(.x = KO1_vs_CTL_sterol_genesets, .f = function(geneset_id){
  gseaPlotESprofile(KO1_vs_CTL_gsea_LIPIDS_results$STEROL, geneSetID = geneset_id, ES_geom="line", NES_table = FALSE, remove_substring = NULL, title = "")
}) %>% setNames(KO1_vs_CTL_sterol_genesets)


# KO2_vs_CTL_gsea_LIPIDS_results
KO2_vs_CTL_LIPID_gseaplots <- purrr::map(.x = lipid_genesets, .f = function(geneset_id){
  gseaPlotESprofile(KO2_vs_CTL_gsea_LIPIDS_results$LIPID, geneSetID = geneset_id, ES_geom="line", NES_table = FALSE, remove_substring = NULL, title = "")
}) %>% setNames(lipid_genesets)

KO2_vs_CTL_STEROL_gseaplots <- purrr::map(.x = KO2_vs_CTL_sterol_genesets, .f = function(geneset_id){
  gseaPlotESprofile(KO2_vs_CTL_gsea_LIPIDS_results$STEROL, geneSetID = geneset_id, ES_geom="line", NES_table = FALSE, remove_substring = NULL, title = "")
}) %>% setNames(KO2_vs_CTL_sterol_genesets)


# saving ggplots ----
# save in a loop
# KO1_vs_CTL_gsea_LIPIDS_results
purrr::map2(.x = names(KO1_vs_CTL_LIPID_gseaplots), .y = "KO1_vs_CTL_LIPID", .f = function(gsea_plot_name, data_name){
  gsea_plot <- KO1_vs_CTL_LIPID_gseaplots[[gsea_plot_name]]

  #ggsave(filename = file.path(results_dir, paste0(data_name,"_", gsea_plot_name,".tiff")), gsea_plot, height = 20, width = 20, units = "cm")
  ggsave(filename = file.path(results_dir, paste0(data_name,"_", gsea_plot_name,".tiff")), gsea_plot)
})

purrr::map2(.x = names(KO1_vs_CTL_STEROL_gseaplots), .y = "KO1_vs_CTL_STEROL", .f = function(gsea_plot_name, data_name){
  gsea_plot <- KO1_vs_CTL_STEROL_gseaplots[[gsea_plot_name]]

  ggsave(filename = file.path(results_dir, paste0(data_name,"_", gsea_plot_name,".tiff")), gsea_plot)
})

# KO2_vs_CTL
purrr::map2(.x = names(KO2_vs_CTL_LIPID_gseaplots), .y = "KO2_vs_CTL_LIPID", .f = function(gsea_plot_name, data_name){
  gsea_plot <- KO2_vs_CTL_LIPID_gseaplots[[gsea_plot_name]]

  #ggsave(filename = file.path(results_dir, paste0(data_name,"_", gsea_plot_name,".tiff")), gsea_plot, height = 20, width = 20, units = "cm")
  ggsave(filename = file.path(results_dir, paste0(data_name,"_", gsea_plot_name,".tiff")), gsea_plot)
})

purrr::map2(.x = names(KO2_vs_CTL_STEROL_gseaplots), .y = "KO2_vs_CTL_STEROL", .f = function(gsea_plot_name, data_name){
  gsea_plot <- KO2_vs_CTL_STEROL_gseaplots[[gsea_plot_name]]

  ggsave(filename = file.path(results_dir, paste0(data_name,"_", gsea_plot_name,".tiff")), gsea_plot)
})

```

## KO1_vs_CTL

Lipid genesets:

```{r KO1_vs_CTL - LIPIDS GSEA results, include=TRUE, eval=TRUE, warning=FALSE}
DT::datatable(dplyr::select(KO1_vs_CTL_gsea_LIPIDS_results_df$LIPID, c("ID", "enrichmentScore", "NES", "pvalue", "p.adjust")) %>% dplyr::arrange(., pvalue),
              rownames = FALSE,
              options = list(pageLength = 10),
              caption = 'Table 1: KO1_vs_CTL - LIPIDS GSEA results.') %>%
  DT::formatRound(c('enrichmentScore', 'NES'), 2) %>%
  DT::formatRound(c('pvalue', 'p.adjust'), 4)
```

```{r KO1_vs_CTL - GOBP_REGULATION_OF_LIPID_METABOLIC_PROCESS,fig.cap = "KO1_vs_CTL: GOBP_REGULATION_OF_LIPID_METABOLIC_PROCESS"}

print(KO1_vs_CTL_LIPID_gseaplots$GOBP_REGULATION_OF_LIPID_METABOLIC_PROCESS)
```

```{r KO1_vs_CTL - GOBP_REGULATION_OF_LIPID_BIOSYNTHETIC_PROCESS,fig.cap = "KO1_vs_CTL: GOBP_REGULATION_OF_LIPID_BIOSYNTHETIC_PROCESS"}

print(KO1_vs_CTL_LIPID_gseaplots$GOBP_REGULATION_OF_LIPID_BIOSYNTHETIC_PROCESS)
```


```{r KO1_vs_CTL - GOBP_LIPID_HOMEOSTASIS,fig.cap = "KO1_vs_CTL: GOBP_LIPID_HOMEOSTASIS"}

print(KO1_vs_CTL_LIPID_gseaplots$GOBP_LIPID_HOMEOSTASIS)
```

Sterol genesets:

```{r KO1_vs_CTL - STEROLS GSEA results, include=TRUE, eval=TRUE, warning=FALSE}
DT::datatable(dplyr::select(KO1_vs_CTL_gsea_LIPIDS_results_df$STEROL, c("ID", "enrichmentScore", "NES", "pvalue", "p.adjust")) %>% dplyr::arrange(., pvalue),
              rownames = FALSE,
              options = list(pageLength = 10),
              caption = 'Table 1: KO1_vs_CTL - STEROLS GSEA results.') %>%
  DT::formatRound(c('enrichmentScore', 'NES'), 2) %>%
  DT::formatRound(c('pvalue', 'p.adjust'), 4)
```

```{r KO1_vs_CTL - GOBP_STEROL_HOMEOSTASIS,fig.cap = "KO1_vs_CTL: GOBP_STEROL_HOMEOSTASIS"}

print(KO1_vs_CTL_STEROL_gseaplots$GOBP_STEROL_HOMEOSTASIS)
```

```{r KO1_vs_CTL - GOBP_REGULATION_OF_CHOLESTEROL_METABOLIC_PROCESS,fig.cap = "KO1_vs_CTL: GOBP_REGULATION_OF_CHOLESTEROL_METABOLIC_PROCESS"}

print(KO1_vs_CTL_STEROL_gseaplots$GOBP_REGULATION_OF_CHOLESTEROL_METABOLIC_PROCESS)
```

## KO2_vs_CTL

Lipid genesets:

```{r KO2_vs_CTL - LIPIDS GSEA results, include=TRUE, eval=TRUE, warning=FALSE}
DT::datatable(dplyr::select(KO2_vs_CTL_gsea_LIPIDS_results_df$LIPID, c("ID", "enrichmentScore", "NES", "pvalue", "p.adjust")) %>% dplyr::arrange(., pvalue),
              rownames = FALSE,
              options = list(pageLength = 10),
              caption = 'Table 1: KO2_vs_CTL - LIPIDS GSEA results.') %>%
  DT::formatRound(c('enrichmentScore', 'NES'), 2) %>%
  DT::formatRound(c('pvalue', 'p.adjust'), 4)
```

```{r KO2_vs_CTL - GOBP_REGULATION_OF_LIPID_METABOLIC_PROCESS,fig.cap = "KO2_vs_CTL: GOBP_REGULATION_OF_LIPID_METABOLIC_PROCESS"}

print(KO2_vs_CTL_LIPID_gseaplots$GOBP_REGULATION_OF_LIPID_METABOLIC_PROCESS)
```

```{r KO2_vs_CTL - GOBP_REGULATION_OF_LIPID_BIOSYNTHETIC_PROCESS,fig.cap = "KO2_vs_CTL: GOBP_REGULATION_OF_LIPID_BIOSYNTHETIC_PROCESS"}

print(KO2_vs_CTL_LIPID_gseaplots$GOBP_REGULATION_OF_LIPID_BIOSYNTHETIC_PROCESS)
```

```{r KO2_vs_CTL - GOBP_LIPID_HOMEOSTASIS,fig.cap = "KO2_vs_CTL: GOBP_LIPID_HOMEOSTASIS"}

print(KO2_vs_CTL_LIPID_gseaplots$GOBP_LIPID_HOMEOSTASIS)
```

Sterol genesets:

```{r KO2_vs_CTL - STEROLS GSEA results, include=TRUE, eval=TRUE, warning=FALSE}
DT::datatable(dplyr::select(KO2_vs_CTL_gsea_LIPIDS_results_df$STEROL, c("ID", "enrichmentScore", "NES", "pvalue", "p.adjust")) %>% dplyr::arrange(., pvalue),
              rownames = FALSE,
              options = list(pageLength = 10),
              caption = 'Table 1: KO2_vs_CTL - STEROLS GSEA results.') %>%
  DT::formatRound(c('enrichmentScore', 'NES'), 2) %>%
  DT::formatRound(c('pvalue', 'p.adjust'), 4)
```

```{r KO2_vs_CTL - GOBP_STEROL_HOMEOSTASIS,fig.cap = "KO2_vs_CTL: GOBP_STEROL_HOMEOSTASIS"}

print(KO2_vs_CTL_STEROL_gseaplots$GOBP_STEROL_HOMEOSTASIS)
```


```{r GSEA results plots - combined, include=FALSE, eval=FALSE}
# combined plots ----
# KO1_vs_CTL_gsea_LIPIDS_results
KO1_vs_CTL_LIPID_comb_gseaplot <- gseaPlotESprofile(KO1_vs_CTL_gsea_LIPIDS_results$LIPID, geneSetID = c("GOBP_REGULATION_OF_LIPID_METABOLIC_PROCESS", "GOBP_REGULATION_OF_LIPID_BIOSYNTHETIC_PROCESS", "GOBP_LIPID_HOMEOSTASIS"), ES_geom="line", NES_table = TRUE, remove_substring = "GOBP_", title = "")

KO1_vs_CTL_STEROL_comb_gseaplot <- gseaPlotESprofile(KO1_vs_CTL_gsea_LIPIDS_results$STEROL, geneSetID = c("GOBP_STEROL_HOMEOSTASIS", "GOBP_REGULATION_OF_CHOLESTEROL_METABOLIC_PROCESS"), ES_geom="line", NES_table = TRUE, remove_substring = "GOBP_", title = "")

# KO2_vs_CTL_gsea_LIPIDS_results
KO2_vs_CTL_LIPID_comb_gseaplot <- gseaPlotESprofile(KO2_vs_CTL_gsea_LIPIDS_results$LIPID, geneSetID = c("GOBP_REGULATION_OF_LIPID_METABOLIC_PROCESS", "GOBP_REGULATION_OF_LIPID_BIOSYNTHETIC_PROCESS"), ES_geom="line", NES_table = TRUE, remove_substring = "GOBP_", title = "")

KO2_vs_CTL_STEROL_comb_gseaplot <- gseaPlotESprofile(KO2_vs_CTL_gsea_LIPIDS_results$STEROL, geneSetID = c("GOBP_STEROL_HOMEOSTASIS"), ES_geom="line", NES_table = TRUE, remove_substring = "GOBP_", title = "")

# saving combined plots
ggsave(filename = file.path(results_dir, paste0("KO1_vs_CTL_LIPID_comb_gseaplot.tiff")), KO1_vs_CTL_LIPID_comb_gseaplot,
       height = 25, width = 25, units = "cm")
ggsave(filename = file.path(results_dir, paste0("KO1_vs_CTL_STEROL_comb_gseaplot.tiff")), KO1_vs_CTL_STEROL_comb_gseaplot,
       height = 25, width = 25, units = "cm")
ggsave(filename = file.path(results_dir, paste0("KO2_vs_CTL_LIPID_comb_gseaplot.tiff")), KO2_vs_CTL_LIPID_comb_gseaplot,
       height = 25, width = 25, units = "cm")
ggsave(filename = file.path(results_dir, paste0("KO2_vs_CTL_STEROL_comb_gseaplot.tiff")), KO2_vs_CTL_STEROL_comb_gseaplot,
       height = 25, width = 25, units = "cm")
```

# sessionInfo
```{r include=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
sessionInfo()

# saving session info
writeLines(capture.output(sessionInfo()), file.path(results_dir, "sessionInfo_them6_gsea.txt"))

renv::snapshot(lockfile = file.path(base_dir, "them6_gsea_renv.lock"))
renv::status(lockfile = file.path(base_dir, "them6_gsea_renv.lock"))
# renv::diagnostics()
```
