---
title: "GSEA analysis - Sevgi (NFATc1)"
author: "Peter Repiscak"
date: '`r format(Sys.time(), "%d %B %Y")`'
output:
 html_document:
  code_folding: hide
  toc: false
  toc_float: false
  theme: united
always_allow_html: true
---

Compare annotation done by Michael Schuster with annotation done here
Check other output files generated!
filtered bam is input file for peak calling!
what is whitelist in the filtered?
check if it makes any difference? filtered and unfiltered?
clean config file and combine RNAseq and ATACseq config files

TODO:
GOBP for WT unique peaks top 25 - simialar plot to the following figures
Check with Florian H. if for the unique peaks we can call the accesible just because they are detected in the WT only
C:\Users\peter.repiscak\OneDrive - CCRI\Bioinformatics\Work\Bionformatics\Projects\Kaan_Boztug\Sevgi_Michael_NFATc1_TMX1\results_dir\figures

TODO:
check and understand the selection of background peak sets!!!
[ ] - for motif analysis we should have used all peaks (to get all motifs); rather than just consensus peaks!!!
[ ] - for FC calculation check https://github.com/GreenleafLab/chromVAR/issues/15

#FIXME
[ ] - heatmap is from deviations not variability - check/correct if wrong label somewhere

Prepare methods + add explanation and description for understanding for deviations and variability
patient renaming:
atacseq
P1180 - P1
P1185 - P2
RJH - HC1
MK - HC2

RNAseq:
double check

ADD:
QC
comparison to previous annotation
methods and references!
description of some of the steps from the pipeline from CeMM
filtering - hg38-ENCODE_whitelist.sorted.v3.bed

https://biomedical-sequencing.at/projects/BSA_0612_NFATC1_T_Cells_ce4d4550ea6744bfbd1046b046aec404/atacseq_report/multiqc_report.html


TO-DO:
replace patient sample names to be consistent with the article naming and with file names when constructing from the raw - e.g. atacseq_P_IKZF2-I325V-Hom
separate analysis into from raw and from processed (as raw requires bam and narrowPeaks!)
clean docker to production ready with renv.lock


1. rewrite library loading
2. rewrite functions into separate file
3. rewrite comments/clean
4. save results files (with checksums)
5.a rewrite workflow into smaller chunks and include or call similar to Mohamed's approach or use targets? metaflow? - create small manageable and iterable modules that are independent (reading and loading files! cache!)
5.b rerun modified and see if we get same results!
6. prepare new renv.lock
7. prepare new docker

```{r setup, eval=FALSE, include=FALSE}
# install packages 
atacseq_packages <- c("bioc::ATACseqQC", "bioc::ChIPseeker", "bioc::ChIPpeakAnno", "bioc::motifmatchr", "bioc::chromVAR", "bioc::ChIPQC", "bioc::MotifDb")
genomic_ranges_packages <- c("bioc::plyranges")
plotting_packages <- c("bioc::ComplexHeatmap", "bioc::karyoploteR", "bioc::tracktables", "ggvenn")
annot_packages <- c("bioc::AnnotationHub", "bioc::biomaRt", "bioc::ensembldb", "bioc::org.Hs.eg.db", "bioc::TxDb.Hsapiens.UCSC.hg38.knownGene")
functional_packages <- c("bioc::clusterProfiler", "bioc::ReactomePA", "bioc::DESeq2", "bioc::JASPAR2020", "bioc::TFBSTools", "bioc::systemPipeR", "bioc::Rsamtools", "bioc::limma", "bioc::GenomicAlignments")
other_packages <- c("here", "openxlsx")

renv::install(atacseq_packages)
renv::install(genomic_ranges_packages)
renv::install(plotting_packages)
renv::install(annot_packages)
renv::install(functional_packages)
renv::install(other_packages)
renv::install("tidyverse")  # check version installed in docker image?
#renv::install("bioc::BSgenome.Hsapiens.UCSC.hg38")
#BiocManager::install("BSgenome.Hsapiens.UCSC.hg38")

#Error in library.dynam(lib, package, package.lib) : 
#  shared object 'openxlsx.so' not found
#renv::install("openxlsx", rebuild=TRUE) # , dependencies=TRUE
renv::install("pheatmap")

base_dir <- "/home/rstudio/workspace/"
renv::snapshot(lockfile = file.path(base_dir, "nfatc1_atacseq_renv.lock"))
renv::status(lockfile = file.path(base_dir, "nfatc1_atacseq_renv.lock"))
```


```{r analysis_parameters, include=FALSE}
# Specifying parameters here ----
if(!("renv" %in% installed.packages())){remotes::install_github("rstudio/renv")}
if(!("import" %in% installed.packages())){renv::install("import")}
if(!("yaml" %in% installed.packages())){renv::install("yaml")}

# rewrite below!
base_dir <- "/home/rstudio/workspace/"
data_dir <- file.path(base_dir, "datasets") 
resources_dir <- file.path(base_dir, "resources")
results_dir <- paste0(base_dir, "results_dir/atacseq")  # file.path
dir.create(results_dir)
dir.create(resources_dir)
setwd(base_dir)

#project setup
# loading config file ----
config_file <- file.path(base_dir, "nfxx_config.yaml")
config <- yaml::read_yaml(config_file)
#config <- yaml::read_yaml(params$config_file)

#project setup
project_name <- config$project_name
nthreads <- 4  # for furrr multisession

# this make take a while...
# https://rstudio.github.io/renv/reference/config.html
# getOption(x, default = NULL)
#renv::settings$use.cache()
# getOption('renv.config.pak.enabled')
# if the project does not automatically activate run:
if(Sys.getenv("RENV_PATHS_CACHE") != "/renv_cache") {Sys.setenv(RENV_PATHS_CACHE = "/renv_cache")}
if(Sys.getenv("RENV_PATHS_LIBRARY") != "/home/rstudio/renv_library") {Sys.setenv(RENV_PATHS_LIBRARY = "/home/rstudio/renv_library")}
#if(Sys.getenv("RENV_CONFIG_PAK_ENABLED") != "TRUE") {Sys.setenv(RENV_CONFIG_PAK_ENABLED = TRUE)} # add pak, targets and benchmarkme to docker!!!
# setting root dir
# 0. renv::activate
renv::activate(project = base_dir)
# DO NOT RUN: 
#renv::init(project = base_dir, bare=TRUE)
# FOR the first run restore environment and packages from them6_gsea_renv.lock 
# 1. restore original environment
renv::restore(project = base_dir, lockfile = file.path(base_dir, "nfatc1_atacseq_renv.lock"), prompt = FALSE)
#file.copy(from = file.path(base_dir, "them6_gsea_renv.lock"), to = file.path(base_dir, "renv.lock")) # to use with renv::diagnostics()
```

```{r run parameters, eval=FALSE, include=FALSE}
#check temp dir
tempdir()

# run parameters ----
# add params:
# add separate document
# define cutoffs and other settings ----
# - [ ] move to params: in the yaml definition
padj_perSample_GOenrich_cutoff <- 0.05
log2FC_cutoff <- log2(1.3) #log2(1.5)
padj_cutoff <- 0.1
motif_padj_cutoff <- 0.05
# within promoters
# using 1000 as in Florian's code
# not using 500 as in the tutorial
promoter_dist <- 1000  # should be the same as above?!
tss_dist <- 1000 # within 500 bp of TSS; or 1000bp?

#report options
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      cache.lazy = FALSE,
                      message=FALSE,
                      warning=FALSE,
                      dev = "png",
                      cache.path = file.path(results_dir,"report","cache/"),
                      fig.path = file.path(results_dir,"report","files/"))
options(width=100)

#renv::snapshot()
```


```{r run in the terminal, eval = FALSE, include = FALSE}
#rmarkdown::render(output_file = stringr::str_replace_all(date(),c(" +"="-", ":"="_")), input = here::here("/home/peter_r/scripts/tiaf1_analysis.Rmd"), output_dir = file.path("/home/peter_r/results/report","html/"))

# rmarkdown::render(input = "/home/peter_r/scripts/tiaf1_analysis.Rmd", output_file = stringr::str_replace_all(date(),c(" +"="-", ":"="_")), output_dir = file.path("/home/peter_r/results/report","html/"))
```

```{r loading libraries, message=FALSE, eval=TRUE, include=FALSE}
import::from(.from = here::here("scripts/utils.R"), "save_rds", "digest_match", .character_only=TRUE)

# Docker needs

# Loading functions from packages ----
# using import to keep clean namespace
# and explicitly specifying in the code (potentially remove)

# annotation
import::from(.from = TxDb.Hsapiens.UCSC.hg38.knownGene, TxDb.Hsapiens.UCSC.hg38.knownGene)
#library("BSgenome.Hsapiens.UCSC.hg38")
import::from(.from = BSgenome.Hsapiens.UCSC.hg38, BSgenome.Hsapiens.UCSC.hg38)
import::from(.from = org.Hs.eg.db, org.Hs.eg.db)

# data processing
import::from(.from = magrittr, "%>%") 
import::from(.from = rtracklayer, import.bed)
import::from(.from = readr, read_csv)
import::from(.from = Rsamtools, BamFileList, ScanBamParam)
import::from(.from = GenomicRanges, GRangesList, reduce)
import::from(.from = RColorBrewer, brewer.pal)
import::from(ggplot2, .all=TRUE) # or just import specific part


# Loading custom functions from atacseq_scripts.R ----
import::from(.from = "atacseq_scripts.R", "remove_blacklisted", "add_flag", "load_peaks_to_GRangesList", "filter_chromosomes_GRangesList") 


# [ ] check if renv works import::from - renv::dependencies()
#  it should still recognize package::method()
# https://rstudio.github.io/renv/reference/dependencies.html

# otherwise using this for renv to discover
## delete
#library("tidyverse")
#library("rtracklayer")
#library("readxl")
#library("GenomicRanges")
#library("RColorBrewer")
#library(BSgenome.Hsapiens.UCSC.hg38)  # cite?!
#library(org.Hs.eg.db)
#dlibrary(ChIPseeker)
#library(ChIPQC)
#library(tracktables)
#library(DT)
#library(DESeq2)
#library(karyoploteR)
#library(seqLogo)
#library(JASPAR2020)
#library(TFBSTools)
#library(ggseqlogo)
#library(motifmatchr)
#library(chromVAR)
#library(BiocParallel)
#BiocManager::install("BSgenome.Hsapiens.NCBI.GRCh38")
# http://bioconductor.org/packages/devel/workflows/vignettes/annotation/inst/doc/Annotation_Resources.html
#library("BSgenome.Hsapiens.NCBI.GRCh38")  # or directly use NCBI fasta
#library(furrr)  # do not load as it masks some of key functions
#class(BSgenome.Hsapiens.UCSC.hg38)
#metadata(BSgenome.Hsapiens.UCSC.hg38)
#BiocManager::install("ATACseqQC", "ChIPseeker", "ChIPpeakAnno", "systemPipeR")
#library("Rsamtools")
#library("ggplot2")
#install.packages("ggupset")
# library(VennDiagram)
#library("magrittr")
#library(ggvenn)
#library("GenomicAlignments")

# ??? are the following used?
#library(MotifDb)
#library(Biostrings)
# library(Rsubread) ? where is this used?
#library(AnnotationHub)
#library(ensembldb)
#library(soGGi)
#library("ATACseqQC")
#library("devtools")



# clean into subsections (as above!)
import::from(.from = ChIPseeker, readPeakFile, covplot, annotatePeak, plotAnnoBar, plotDistToTSS, plotAnnoPie, upsetplot, as.GRanges)
import::from(.from = ChIPQC, ChIPQC, flagtagcounts, plotSSD, plotCC, plotRegi, ChIPQCreport)
import::from(.from = DESeq2, DESeqDataSetFromMatrix, estimateSizeFactors, sizeFactors, counts, DESeq, rlog, plotPCA, plotDispEsts, results)
import::from(.from = karyoploteR, makeGenesDataFromTxDb, plotKaryotype, addGeneNames, mergeTranscripts, kpPlotGenes)
import::from(.from = JASPAR2020, JASPAR2020)
import::from(.from = TFBSTools, getMatrixSet)
import::from(.from = BiocParallel, register, MulticoreParam, SerialParam)
import::from(.from = systemPipeR, overLapper, vennPlot, intersectmatrix)
import::from(.from = Rsamtools, BamFileList, ScanBamParam, FaFile)
import::from(.from = limma, vennCounts, vennDiagram)
import::from(.from = ggvenn, ggvenn)
import::from(.from = GenomicAlignments, summarizeOverlaps)
import::from(.from = clusterProfiler, enrichGO)
import::from(.from = GenomeInfoDb, seqlevels, "seqlevels<-", "seqlevelsStyle<-", seqnames, keepSeqlevels, seqlengths)  # seqlevels<- needed in ChIPQC
import::from(.from = SummarizedExperiment, rowRanges, assay)


# update docker!

#saving session info
#sink(paste0(results_dir, "/session_info.txt"), append = FALSE)
# print() needed to work when used with 'Source on Save' in RStudio
# without it, sink() will save an empty file
# print(sessioninfo::session_info())
#sink()

# xfun::session_info()
```


```{r loading annotations, message=FALSE, eval=TRUE, include=TRUE}
# defining main chromosomes ----
mainChromosomes <- paste0("chr", c(1:22, "X", "Y"))  # , "M"


# loading txdb ----
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

# gencode
#https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_40/gencode.v40.annotation.gtf.gz

# txdb <- GenomicFeatures::makeTxDbFromEnsembl(organism="Homo sapiens",
#                    release=106,
#                    circ_seqs=NULL,
#                    server="ensembldb.ensembl.org",
#                    username="anonymous", password=NULL, port=0L,
#                    tx_attrib=NULL)

print(txdb)
#TxDb object:
# Db type: TxDb
# Supporting package: GenomicFeatures
# Data source: UCSC
# Genome: hg38
# Organism: Homo sapiens
# Taxonomy ID: 9606
# UCSC Table: knownGene
# UCSC Track: GENCODE V39
# Resource URL: http://genome.ucsc.edu/
# Type of Gene ID: Entrez Gene ID
# Full dataset: yes
# miRBase build ID: NA
# Nb of transcripts: 266064
# Db created by: GenomicFeatures package from Bioconductor
# Creation time: 2022-04-07 16:50:30 +0000 (Thu, 07 Apr 2022)
# GenomicFeatures version at creation time: 1.47.13
# RSQLite version at creation time: 2.2.11
# DBSCHEMAVERSION: 1.2


# loading blacklisted regions ----
# make available offline
# simplify checks below
backlist_filename <- "hg38.blacklist.bed.gz"
backlist_filename_md5sum <- "4a58b07c400e8641dc3df79ad6fcf12c"
blacklist_file_url <- paste0("http://mitra.stanford.edu/kundaje/akundaje/release/blacklists/hg38-human/", backlist_filename)

if (!file.exists(file.path(resources_dir, backlist_filename))) {
  
  message("downloading ", backlist_filename)
  download.file(url = blacklist_file_url,
                destfile = file.path(resources_dir, backlist_filename))
  
  # creating md5sum for file with blacklisted regions
  message("generating md5sum for ", backlist_filename)
  backlist_file_conn <- file(file.path(resources_dir, paste0(backlist_filename, ".md5sum")))
  writeLines(paste(as.vector(tools::md5sum(file.path(resources_dir, backlist_filename))), backlist_filename, sep = " "), backlist_file_conn)
  close(backlist_file_conn)
  
} else if (as.vector(tools::md5sum(file.path(resources_dir, backlist_filename))) == backlist_filename_md5sum){
  message("md5sum is OK for the existing ", backlist_filename)
  blacklist <- rtracklayer::import.bed(file.path(resources_dir, backlist_filename))
} else {
  message("Something weird happened!")
  message("md5sum differs for the downloaded ", backlist_filename)
  message("md5sum should be: ", backlist_filename_md5sum)
}


# online download
#blacklist_test <- rtracklayer::import.bed("http://mitra.stanford.edu/kundaje/akundaje/release/blacklists/hg38-human/hg38.blacklist.bed.gz")

#identical(blacklist, blacklist_test)  # TRUE


```

# Initialization
### Loading data and metadata

```{r loading data, include=TRUE, eval=TRUE}
# Loading atac-seq data ----

#TODO:
# [ ] - add md5sums (ubam, narrowPeaks, bam files, bigwig) of files and metadata file!? - nfxx_bamfiles.md5sums

# loading metadata from excel file
project_metadata_raw <- readr::read_csv(file = file.path(base_dir, config$data_files$atacseq$metadata))
 
# loading peaks files ----
narrowPeak_files <- list.files(path = data_dir, pattern = "*_peaks.narrowPeak$", full.names=TRUE, recursive=TRUE)
narrowPeak_df <- data.frame(narrowPeak_files_path = narrowPeak_files) %>%
  dplyr::mutate(sample_name = gsub(pattern = "(.+/)(.+)(_peaks.narrowPeak)", replacement = "\\2", narrowPeak_files),
                narrowPeak_files = gsub(pattern = "(.+/)(.+_peaks.narrowPeak)", replacement = "\\2", narrowPeak_files))

# loading bam files ----
bam_files <- list.files(path = data_dir, pattern = "*.filtered.bam$", full.names=TRUE, recursive=TRUE)
bam_df <- data.frame(bam_files_path = bam_files) %>%
  dplyr::mutate(sample_name = gsub(pattern = "(.+/)(.+)(.filtered.bam)", replacement = "\\2", bam_files),
                bam_files = gsub(pattern = "(.+/)(.+.filtered.bam)", replacement = "\\2", bam_files))

# updating metadata wiht peak and bam files information
project_metadata <- project_metadata_raw %>%
  dplyr::left_join(., narrowPeak_df, by = "sample_name") %>%
  dplyr::left_join(., bam_df, by = "sample_name") 

rownames(project_metadata) <- project_metadata$sample_name
```

### Quality control analysis

```{r QC analysis, include=TRUE, eval=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
# ADD description

sample_names <- project_metadata$sample_name

peak_files <- project_metadata$narrowPeak_files_path
names(peak_files) <- project_metadata$sample_name

# Loading peaks ----
peaks_grlist_rds <- file.path(results_dir, paste0(project_name, "_peaks_grlist.rds"))
peaks_grlist_rds_digest <- file.path(results_dir, paste0(project_name, "_rnk_lists_digest.md5"))
peaks_grlist_rds_digest_md5sum <- "8ae4e32277711427863c6844cc20002c"  # to fix for final results otherwise source from _digest.md5 file

if (!file.exists(peaks_grlist_rds)) {
  message("Loading peaks into peaks_grlist...")
  peak_gr_raw <- load_peaks_to_GRangesList(peak_files)
  
  # keeping only main chromosomes ----
  # remove non-standard chromosomes and mitochondrial!
  peak_grlist <- filter_chromosomes_GRangesList(granges_list = peak_gr_raw)
  
  # saving
  save_rds(x = peak_grlist,
           rdsname = peaks_grlist_rds, 
           digestname = peaks_grlist_rds_digest)
  
} else {
  message("peaks_grlist exists...loading!")
  
  # md5_digest them6_gsea_results.rds: 707217f5b76a5db6a36526ac6fce78c9
  peak_grlist <- readRDS(peaks_grlist_rds) 
  peak_grlists_md5 <- digest::digest(peaks_grlist_rds_digest, algo = "md5")

  # check if correct data is loaded in case of missing _digest.md5
  if (digest_match(x = peak_grlist, digestfile = peaks_grlist_rds_digest)) {
    # checking for match with _digest.md5
      message("loaded data md5 matches md5 in the digest file")
  }  
  
  # check if md5 corresponds to the final (production) md5
  if (!is.null(peaks_grlist_rds_digest_md5sum) & peak_grlists_md5 != peaks_grlist_rds_digest_md5sum) {
    warning("md5 is: ", peak_grlists_md5, " and it differs from expected production file md5: ", peaks_grlist_rds_digest_md5sum)
  }
}

# Performing QC analysis ----
# TODO:
# [ ] - rewrite QC into functions and refactor this section
# [ ] - add targets pipeline

qc_rds <- file.path(results_dir, paste0(project_name, "_QCresults.rds"))
qc_rds_digest <- file.path(results_dir, paste0(project_name, "_QCresults_digest.md5"))
qc_rds_digest_md5sum <- "8ae4e32277711427863c6844cc20002c"  # to fix for final results otherwise source from _digest.md5 file
if (!file.exists(qc_rds)){
  # coverage over chromosomes
  # also use GRangesList to calculate per sample
  # https://guangchuangyu.github.io/2016/02/covplot-supports-grangeslist/
  peak_gr_per_chr <- ChIPseeker::covplot(peak = peak_grlist,
                                         title ="ATAC-seq peaks coverage over chromosomes")
  
  #! store
  peaks_per_chr_plot <- peak_gr_per_chr + facet_grid(chr ~ .id)  # remove non-main chromosomes
  
  # looking at annotated regions 
  peak_grlist_annot <- lapply(peak_grlist, ChIPseeker::annotatePeak, 
                         TxDb=txdb,
                         tssRegion=c(-3000, 3000),
                         level = "gene",  
                         annoDb="org.Hs.eg.db", 
                         verbose=FALSE)
  
  peak_grlist_annot_bar_plot <- ChIPseeker::plotAnnoBar(peak_grlist_annot)
  peak_grlist_annot_distTSS_plot <- ChIPseeker::plotDistToTSS(peak_grlist_annot,
              title="Distribution of ATAC-seq peaks \nrelative to TSS")
  
  ggsave(peak_grlist_annot_bar_plot, filename = file.path(results_dir, "/peak_grlist_annot_bar_plot.png"))
  ggsave(peak_grlist_annot_bar_plot, filename = file.path(results_dir, "/peak_grlist_annot_bar_plot.pdf"))
  ggsave(peak_grlist_annot_distTSS_plot, filename = file.path(results_dir, "/peak_grlist_annot_distTSS_plot.png"))
   
  # constructing ChIPQCexperiment object
  atacseq_experiment_df <- project_metadata %>%
    dplyr::select(sample_name, genotype, bam_files_path, narrowPeak_files_path) %>%
    dplyr::rename(SampleID = sample_name, 
                  Condition = genotype,
                  bamReads = bam_files_path,
                  Peaks = narrowPeak_files_path) %>%
    dplyr::mutate(PeakCaller = "narrow")
  
  # consensus QC ----
  # quite computationally expensive so save in a separate .RData
  # TODO:
  # run also ChIPQC in consensus mode!
  #library(ChIPQC) # !!! use import
  # FIXME:
  # potential issue with bam files?! https://support.bioconductor.org/p/112963/
  # potential issue with blacklist? https://support.bioconductor.org/p/97738/
  atacseq_experiment_df_filt <- atacseq_experiment_df %>%
    dplyr::mutate(Factor=Condition,
                  bamReads = gsub(pattern="\\.filtered", replacement = "", bamReads))
  blacklist_ensembl <- blacklist
  seqlevelsStyle(blacklist_ensembl) <- "Ensembl"
  atacseqQC <- ChIPQC::ChIPQC(atacseq_experiment_df_filt,
                              annotation="hg38",
                              chromosomes = "chr1",  # mainChromosomes
                              consensus=TRUE, # TRUE?
                              blacklist = blacklist_ensembl)
  
  atacseqQC_metrics <- as.data.frame(QCmetrics(atacseqQC))
  atacseqQC_SSDplot <- ChIPQC::plotSSD(atacseqQC) + xlim(0.13, 0.18)  # Standardized Standard Deviation
  # SSD is a measure of standard deviation of signal across the genome with higher scores reflecting significant pile-up of reads. SSD can therefore be used to assess both the extent of ultra high signals and the signal
  atacseqQC_CCplot <- ChIPQC::plotCC(atacseqQC, facetBy = "Sample")  # Cross-coverage plot
  
  ggsave(atacseqQC_SSDplot, filename = file.path(results_dir, "/atacseqQC_SSDplot.png"))
  ggsave(atacseqQC_CCplot, filename = file.path(results_dir, "/atacseqQC_CCplot.png"))
  
  # duplication rate
  atacseqQC_flags <- ChIPQC::flagtagcounts(atacseqQC)
  #atacseqQC_flags["DuplicateByChIPQC", ]/atacseqQC_flags["Mapped", ]
  
  # review distribution of reads across features
  atacseqQC_feature_dist <- ChIPQC::plotRegi(atacseqQC) + coord_flip() + theme_bw()
  ggsave(atacseqQC_feature_dist, filename = file.path(results_dir, "/atacseqQC_feature_dist.png"))
  
  # add to report
  ChIPQC::ChIPQCreport(object = atacseqQC, 
                       reportName="ATACseq_QCreport",
                       #facetBy=c("sample_group"),
                       reportFolder = paste0(results_dir, "/ATACseq_QCreport"))
  
  message("Saving QC_results")
  QCresults <- list(peaks_per_chr_plot=peaks_per_chr_plot, 
                    atacseqQC=atacseqQC, 
                    atacseqQC_metrics=atacseqQC_metrics, 
                    atacseqQC_SSDplot=atacseqQC_SSDplot, 
                    atacseqQC_flags=atacseqQC_flags)
  
  save_rds(x = QCresults,
           rdsname = qc_rds, 
           digestname = qc_rds_digest)
  
} else {
  message("Loading QC_results")
  
  # md5_digest them6_gsea_results.rds: 707217f5b76a5db6a36526ac6fce78c9
  QCresults <- readRDS(qc_rds) 
  QCresults_md5 <- digest::digest(qc_rds_digest, algo = "md5")

  peaks_per_chr_plot <- QCresults$peaks_per_chr_plot
  atacseqQC <- QCresults$atacseqQC 
  atacseqQC_metrics <- QCresults$atacseqQC_metrics
  atacseqQC_SSDplot <- QCresults$atacseqQC_SSDplot
  atacseqQC_flags <- QCresults$atacseqQC_flags
  
  # check if correct data is loaded in case of missing _digest.md5
  if (digest_match(x = QCresults, digestfile = qc_rds_digest)) {
    # checking for match with _digest.md5
      message("loaded data md5 matches md5 in the digest file")
  }  
  
  # check if md5 corresponds to the final (production) md5
  if (!is.null(qc_rds_digest_md5sum) & QCresults_md5 != qc_rds_digest_md5sum) {
    warning("md5 is: ", QCresults_md5, " and it differs from expected production file md5: ", qc_rds_digest_md5sum)
  }
}


# Pre-processing ----
# removing blacklist regions per sample
# removing blacklist regions per sample set:
#    to generate universe of genes for enrichment?! 
#    for downstream differential accessibility analysis
# check if remove blacklist per sample = remove blacklist per combined

if (!file.exists(paste0(results_dir, "/peak_grlist_filt.RDS"))){
  
  peak_gr_filt_list <- purrr::map(.x = sample_names, .f = function(sample_name) {
    message("Removing blaclisted regions from: ", sample_name)

    remove_blacklisted(peak_gr = peak_grlist[[sample_name]])
    
  }) %>% setNames(sample_names)
  
  peak_grlist_filt <- GenomicRanges::GRangesList(peak_gr_filt_list)
  message("Saving peak_grlist_filt")
  saveRDS(peak_grlist_filt, file = paste0(results_dir, "/peak_grlist_filt.RDS"))
  
} else {
  message("Loading peak_grlist_filt")
  peak_grlist_filt <- readRDS(file = paste0(results_dir, "/peak_grlist_filt.RDS"))
}

# Annotating peaks ----

if (!file.exists(paste0(results_dir, "/allPeaksSet_filt_nR_annot.RData"))){
  allPeaksSet_filt_nR <- GenomicRanges::reduce(unlist(peak_grlist_filt)) 
  peaks_overlap <- list()
  for (i in 1:length(peak_grlist_filt)) {
    peaks_overlap[[i]] <-  IRanges::overlapsAny(allPeaksSet_filt_nR, peak_grlist_filt[[i]])
  }
  overlapMatrix <- do.call(cbind, peaks_overlap)
  colnames(overlapMatrix) <- names(peak_grlist_filt)  # project_metadata$sample_name
  mcols(allPeaksSet_filt_nR) <- overlapMatrix
  
  # annotating to get gene universe for enrichment ----
  # txdb_ah_ensembl, txdb
  length(allPeaksSet_filt_nR)
  # annotating to gene!
  # how does results differ when annotating to gene or transcript?!
  allPeaksSet_filt_nR_annot <- ChIPseeker::annotatePeak(peak = allPeaksSet_filt_nR,
                                                        tssRegion=c(-3000, 3000),
                                                        TxDb=txdb, 
                                                        level = "gene",  # "transcript" or "gene"; default is transcript
                                                        annoDb="org.Hs.eg.db")
  
  allPeaksSet_filt_nR_annot_gr <- as.GRanges(allPeaksSet_filt_nR_annot)
  # defining unique identifier as a combination of entrezID_chr_start_end
  allPeaksSet_filt_nR_annot_gr$unique_geneID <- paste(allPeaksSet_filt_nR_annot_gr$geneId, as.character(seqnames(allPeaksSet_filt_nR_annot_gr)), IRanges::start(allPeaksSet_filt_nR_annot_gr), IRanges::end(allPeaksSet_filt_nR_annot_gr), sep = "_")
  message("Duplicated unique_geneID: ", sum(duplicated(allPeaksSet_filt_nR_annot_gr$unique_geneID)))
  allPeaksSet_filt_nR_annot_df <- as.data.frame(allPeaksSet_filt_nR_annot_gr)
  
  message("Saving allPeaksSet_filt_nR_annot")
  save(allPeaksSet_filt_nR, allPeaksSet_filt_nR_annot, allPeaksSet_filt_nR_annot_gr, allPeaksSet_filt_nR_annot_df, file = paste0(results_dir, "/allPeaksSet_filt_nR_annot.RData"))
} else {
  message("Loading allPeaksSet_filt_nR_annot")
  load(file = paste0(results_dir, "/allPeaksSet_filt_nR_annot.RData"))
}

# assesing distanceToTSS of annotated features 
allPeaksSet_filt_nR_annot_df_perGroup <- allPeaksSet_filt_nR_annot_df %>%
  dplyr::select(PID1185, PID1180, ND_RJH, ND_MK, distanceToTSS) %>%
  tidyr::pivot_longer(., -distanceToTSS, names_to = "sample_name", values_to = "detected") %>%
  dplyr::filter(detected == TRUE)

sum(abs(allPeaksSet_filt_nR_annot_df_perGroup$distanceToTSS) > 50000)/length(allPeaksSet_filt_nR_annot_df_perGroup$distanceToTSS)
# there are few extreme distances to TSS - investigate

#! store
annot_pie_plot <- ChIPseeker::plotAnnoPie(allPeaksSet_filt_nR_annot)
annot_bar_plot <- ChIPseeker::plotAnnoBar(allPeaksSet_filt_nR_annot)

annot_distTSS_plot <- ChIPseeker::plotDistToTSS(allPeaksSet_filt_nR_annot,
              title="Distribution of ATAC-seq peaks \nrelative to TSS")

ggsave(annot_distTSS_plot, filename = file.path(results_dir, "/annot_distTSS_plot.png"))
ggsave(annot_bar_plot, filename = file.path(results_dir, "/annot_bar_plot.png"))

png(filename = file.path(results_dir, "/annot_pie_plot.png"))
ChIPseeker::plotAnnoPie(allPeaksSet_filt_nR_annot)
dev.off()

pdf(file = file.path(results_dir, "/annot_pie_plot.pdf"))
ChIPseeker::plotAnnoPie(allPeaksSet_filt_nR_annot)
dev.off()

# defining gene universe used for enrichment 
gene_universe <- unique(allPeaksSet_filt_nR_annot_df$geneId)
length(gene_universe)

# Qualitative analysis: per sample  ----
# subsetting peaks to those around TSS regions (+/- 1000) for functional analysis

venncounts_object <- limma::vennCounts(mcols(allPeaksSet_filt_nR))
limma::vennDiagram(venncounts_object, main="Overlapping peaks")

png(filename = file.path(results_dir, "/overlapping_peaks.png"))
limma::vennDiagram(venncounts_object, main="Overlapping peaks")
dev.off()

# FIXME:
# tibble::as_tibble(mcols(allPeaksSet_filt_nR))
peak_overlaps_df <- tibble::as_tibble(mcols(allPeaksSet_filt_nR)) %>%
  dplyr::mutate(row_sums = rowSums(.)) %>%
  tibble::rowid_to_column(var="rowid") %>%
  dplyr::mutate(value=paste0("peak", rowid)) %>%
  dplyr::select(rowid, value, PID1185, PID1180, ND_RJH, ND_MK) %>%
  dplyr::mutate(PID1185 = ifelse(PID1185, paste0("peak",rowid), NA),
                PID1180 = ifelse(PID1180, paste0("peak",rowid), NA),
                ND_RJH = ifelse(ND_RJH, paste0("peak",rowid), NA),
                ND_MK = ifelse(ND_MK, paste0("peak",rowid), NA))

#TODO:
# [ ] - check why this differ from the limma venn?
peak_overlaps_list <- list(PID1185=peak_overlaps_df$PID1185[!is.na(peak_overlaps_df$PID1185)],
                 PID1180=peak_overlaps_df$PID1180[!is.na(peak_overlaps_df$PID1180)],
                 ND_RJH=peak_overlaps_df$ND_RJH[!is.na(peak_overlaps_df$ND_RJH)],
                 ND_MK=peak_overlaps_df$ND_MK[!is.na(peak_overlaps_df$ND_MK)])
lapply(peak_overlaps_list, length)

peaks_venn <- ggvenn::ggvenn(
  peak_overlaps_list, 
  #fill_color = c("#0073C2FF", "#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5) + labs(title ="Overlapping ATAC-seq peaks")

ggsave(peaks_venn, filename = file.path(results_dir, "/overlapping_peaks.png"))
ggsave(peaks_venn, filename = file.path(results_dir, "/overlapping_peaks.pdf"))

```

### Per sample analysis

Analysis of peaks unique to each of the samples.

```{r perSample analysis, include=TRUE, eval=TRUE, cache=TRUE}
# creating combined reduced object:
#   used to define gene universe for enrichment analysis
#   used to find consensus peaks

#occurrences <- rowSums(as.data.frame(elementMetadata(nrToCount)))
# in two and more samples!
#consensusToCount <- nrToCount[occurrences >= 2, ]  # potentially ignore Clone?!
unique_occurences <- rowSums(as.data.frame(elementMetadata(allPeaksSet_filt_nR))) == 1
gene_universe <- unique(allPeaksSet_filt_nR_annot_df$geneId)
length(gene_universe)

#TODO:
# [ ] - add per mut vs germline
# [ ] - save results of this enrichment!!!

# venncounts_object <- limma::vennCounts(mcols(allPeaksSet_filt_nR_annot_gr))
# limma::vennDiagram(venncounts_object, main="Overlapping peaks")
#peaks_occurences <- rowSums(allPeaksSet_filt_nR_annot_gr$ND_RJH)
NFATc1_mut_only_anyP <- (allPeaksSet_filt_nR_annot_gr$ND_RJH == FALSE) & (allPeaksSet_filt_nR_annot_gr$ND_MK == FALSE)
NFATc1_mut_only_bothP <- (allPeaksSet_filt_nR_annot_gr$ND_RJH == FALSE) & (allPeaksSet_filt_nR_annot_gr$ND_MK == FALSE) & (allPeaksSet_filt_nR_annot_gr$PID1185 == TRUE) & (allPeaksSet_filt_nR_annot_gr$PID1180 == TRUE)
WT_only_anyP <- (allPeaksSet_filt_nR_annot_gr$PID1185 == FALSE) & (allPeaksSet_filt_nR_annot_gr$PID1180 == FALSE)

peaks_NFATc1_mut_only <- as.data.frame(allPeaksSet_filt_nR_annot_gr[NFATc1_mut_only_anyP]) %>%  
  dplyr::mutate(peakOnly = "NFATc1_mut")

peaks_NFATc1_mut_only_bothP <- as.data.frame(allPeaksSet_filt_nR_annot_gr[NFATc1_mut_only_bothP]) %>%  
  dplyr::mutate(peakOnly = "NFATc1_mut_bothP")

peaks_WT_only <- as.data.frame(allPeaksSet_filt_nR_annot_gr[WT_only_anyP]) %>%  
  dplyr::mutate(peakOnly = "WT")

# Only in patient
# peaks_PID1185_only <- as.data.frame(allPeaksSet_filt_nR_annot_gr[(allPeaksSet_filt_nR_annot_gr$PID1185 == TRUE) & unique_occurences]) %>%  
#   dplyr::mutate(peakOnly = "PID1185")
# 
# peaks_PID1180_only <- as.data.frame(allPeaksSet_filt_nR_annot_gr[(allPeaksSet_filt_nR_annot_gr$PID1180 == TRUE) & unique_occurences]) %>%  
#   dplyr::mutate(peakOnly = "PID1180")

# Only in ND1
# 
# peaks_ND_RJH_only <- as.data.frame(allPeaksSet_filt_nR_annot_gr[(allPeaksSet_filt_nR_annot_gr$ND_RJH == TRUE) & unique_occurences]) %>%  
#   dplyr::mutate(peakOnly = "ND_RJH")
# 
# # Only in ND2
# peaks_ND_MK_only <- as.data.frame(allPeaksSet_filt_nR_annot_gr[(allPeaksSet_filt_nR_annot_gr$ND_MK == TRUE) & unique_occurences]) %>%  
#   dplyr::mutate(peakOnly = "ND_MK")

# enriching for genes found in unique peaks ----
#genes_overlap <- as.data.frame(allPeaksSet_filt_nR_annot_gr) do all at once!
# genes_overlap <- dplyr::bind_rows(peaks_PID1185_only, peaks_PID1180_only, peaks_ND_RJH_only, peaks_ND_MK_only) 
# table(genes_overlap$peakOnly)  
genes_overlap <- dplyr::bind_rows(peaks_NFATc1_mut_only, peaks_WT_only) 
genes_overlap_bothP <- dplyr::bind_rows(peaks_NFATc1_mut_only_bothP, peaks_WT_only) 
table(genes_overlap$peakOnly)  
table(genes_overlap_bothP$peakOnly)  

genes_overlap_toEnrich_df <- genes_overlap %>%
  dplyr::group_by(peakOnly) %>%
  dplyr::summarise(unique_genes = list(unique(geneId)))

genes_overlap_toEnrich_df_bothP <- genes_overlap_bothP %>%
  dplyr::group_by(peakOnly) %>%
  dplyr::summarise(unique_genes = list(unique(geneId)))

lapply(genes_overlap_toEnrich_df$unique_genes, length)
lapply(genes_overlap_toEnrich_df_bothP$unique_genes, length)

genes_overlap_toEnrich <- genes_overlap_toEnrich_df$unique_genes
names(genes_overlap_toEnrich) <- genes_overlap_toEnrich_df$peakOnly

genes_overlap_toEnrich_bothP <- genes_overlap_toEnrich_df_bothP$unique_genes
names(genes_overlap_toEnrich_bothP) <- genes_overlap_toEnrich_df_bothP$peakOnly

# identifying patient unique genes within N bp from TSS
genes_overlap_toEnrich_vennset <- systemPipeR::overLapper(genes_overlap_toEnrich, type="vennsets")
genes_overlap_toEnrich_intersect <- systemPipeR::overLapper(genes_overlap_toEnrich, type="intersects")
vennplot_example <- systemPipeR::vennPlot(genes_overlap_toEnrich_vennset, mymain="Overlapping genes between sample specific peaks")

# bothP
genes_overlap_toEnrich_vennset_bothP <- systemPipeR::overLapper(genes_overlap_toEnrich_bothP, type="vennsets")
genes_overlap_toEnrich_intersect_bothP <- systemPipeR::overLapper(genes_overlap_toEnrich_bothP, type="intersects")
vennplot_example_bothP <- systemPipeR::vennPlot(genes_overlap_toEnrich_vennset_bothP, mymain="Overlapping genes between sample specific peaks")

genes_overlap_toEnrich_intersect_df <- as.data.frame(systemPipeR::intersectmatrix(genes_overlap_toEnrich_vennset)) %>%
  dplyr::mutate(overlaps = rowSums(.)) %>%
  tibble::rownames_to_column(var = "geneId") %>%
  #dplyr::filter(overlaps == 1) %>% not filtering for unique genes otherwise not enough significant enriching
  dplyr::select(-overlaps) %>%
  tidyr::pivot_longer(., -geneId, names_to = "sample_name", values_to = "geneOnly") %>%
  dplyr::filter(geneOnly == 1) %>%
  dplyr::group_by(sample_name) %>%
  dplyr::summarise(unique_genes = list(geneId))

# bothP
genes_overlap_toEnrich_intersect_df_bothP <- as.data.frame(systemPipeR::intersectmatrix(genes_overlap_toEnrich_vennset_bothP)) %>%
  dplyr::mutate(overlaps = rowSums(.)) %>%
  tibble::rownames_to_column(var = "geneId") %>%
  #dplyr::filter(overlaps == 1) %>% not filtering for unique genes otherwise not enough significant enriching
  dplyr::select(-overlaps) %>%
  tidyr::pivot_longer(., -geneId, names_to = "sample_name", values_to = "geneOnly") %>%
  dplyr::filter(geneOnly == 1) %>%
  dplyr::group_by(sample_name) %>%
  dplyr::summarise(unique_genes = list(geneId))

unique_genes_toEnrich <- genes_overlap_toEnrich_intersect_df$unique_genes
names(unique_genes_toEnrich) <- genes_overlap_toEnrich_intersect_df$sample_name
lapply(unique_genes_toEnrich, length)

# bothP
unique_genes_toEnrich_bothP <- genes_overlap_toEnrich_intersect_df_bothP$unique_genes
names(unique_genes_toEnrich_bothP) <- genes_overlap_toEnrich_intersect_df_bothP$sample_name
lapply(unique_genes_toEnrich_bothP, length)

# enriching for GO-BP
# not a lot of significant go terms?! -> enriching first then filtering on go terms!?
enrichment_names <- names(unique_genes_toEnrich)
perSample_GOBP_enrich_list <- purrr::map(.x = names(unique_genes_toEnrich), .f = function(sample_name) {
  message("Enriching GOBP for unique genes in: ", sample_name)
  
  clusterProfiler::enrichGO(gene = unique_genes_toEnrich[[sample_name]],
                universe      = gene_universe,
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",  # MF, CC, BP, ALL
                pAdjustMethod = "BH", 
                pvalueCutoff  = 1,
                qvalueCutoff  = 1,
                readable      = TRUE)
  
}) %>% setNames(names(unique_genes_toEnrich))

# bothP
enrichment_names_bothP <- names(unique_genes_toEnrich_bothP)
perSample_GOBP_enrich_list_bothP <- purrr::map(.x = names(unique_genes_toEnrich_bothP), .f = function(sample_name) {
  message("Enriching GOBP for unique genes in: ", sample_name)
  
  clusterProfiler::enrichGO(gene = unique_genes_toEnrich_bothP[[sample_name]],
                universe      = gene_universe,
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",  # MF, CC, BP, ALL
                pAdjustMethod = "BH", 
                pvalueCutoff  = 1,
                qvalueCutoff  = 1,
                readable      = TRUE)
  
}) %>% setNames(names(unique_genes_toEnrich_bothP))

# identify significant go terms
perSample_GOBP_enrich_df <- purrr::map(.x=perSample_GOBP_enrich_list, .f = function(enrich_res){
  as.data.frame(enrich_res) 
}) %>% setNames(names(perSample_GOBP_enrich_list))

# bothP
perSample_GOBP_enrich_df_bothP <- purrr::map(.x=perSample_GOBP_enrich_list_bothP, .f = function(enrich_res){
  as.data.frame(enrich_res) 
}) %>% setNames(names(perSample_GOBP_enrich_list_bothP))

perSample_GOBP_enrich_df_signif <- purrr::map(.x=perSample_GOBP_enrich_list, .f = function(enrich_res){
  enrich_res_df_signif <- as.data.frame(enrich_res) %>% 
    dplyr::filter(p.adjust < padj_perSample_GOenrich_cutoff)
  return(enrich_res_df_signif)
}) %>% setNames(names(perSample_GOBP_enrich_list))

# bothP
perSample_GOBP_enrich_df_signif_bothP <- purrr::map(.x=perSample_GOBP_enrich_list_bothP, .f = function(enrich_res){
  enrich_res_df_signif <- as.data.frame(enrich_res) %>% 
    dplyr::filter(p.adjust < padj_perSample_GOenrich_cutoff)
  return(enrich_res_df_signif)
}) %>% setNames(names(perSample_GOBP_enrich_list_bothP))

perSample_GOBP_enrich_df_signif_ID <- purrr::map(.x=perSample_GOBP_enrich_df_signif, .f = function(enrich_res){
    dplyr::pull(enrich_res, ID)
}) %>% setNames(names(perSample_GOBP_enrich_df_signif))

# bothP
perSample_GOBP_enrich_df_signif_ID_bothP <- purrr::map(.x=perSample_GOBP_enrich_df_signif_bothP, .f = function(enrich_res){
    dplyr::pull(enrich_res, ID)
}) %>% setNames(names(perSample_GOBP_enrich_df_signif_bothP))

lapply(perSample_GOBP_enrich_df_signif_ID, length)
lapply(perSample_GOBP_enrich_df_signif_ID_bothP, length)

perSample_GOBP_enrich_signif_vennset <- systemPipeR::overLapper(perSample_GOBP_enrich_df_signif_ID, type="vennsets")
perSample_GOBP_enrich_signif_intersect <- systemPipeR::intersectmatrix(systemPipeR::overLapper(perSample_GOBP_enrich_df_signif_ID, type="intersects"))
systemPipeR::vennPlot(perSample_GOBP_enrich_signif_vennset, mymain="Overlapping GOBP terms between sample specific peaks", type="ellipse")

# bothP
perSample_GOBP_enrich_signif_vennset_bothP <- systemPipeR::overLapper(perSample_GOBP_enrich_df_signif_ID_bothP, type="vennsets")
perSample_GOBP_enrich_signif_intersect_bothP <- systemPipeR::intersectmatrix(systemPipeR::overLapper(perSample_GOBP_enrich_df_signif_ID_bothP, type="intersects"))
systemPipeR::vennPlot(perSample_GOBP_enrich_signif_vennset_bothP, mymain="Overlapping GOBP terms between sample specific peaks", type="ellipse")

png(filename = file.path(results_dir, "/perSample_GOBP_enrich_df_signif_ID.png"))
systemPipeR::vennPlot(perSample_GOBP_enrich_signif_vennset, mymain="Overlapping GOBP terms between sample specific peaks", type="ellipse")
dev.off()

# bothP
png(filename = file.path(results_dir, "/perSample_bothP_GOBP_enrich_df_signif_ID.png"))
systemPipeR::vennPlot(perSample_GOBP_enrich_signif_vennset_bothP, mymain="Overlapping GOBP terms between sample specific peaks - bothP", type="ellipse")
dev.off()

perSample_GOBP_enrich_df_signif_collapse <- dplyr::bind_rows(perSample_GOBP_enrich_df_signif) %>%
  dplyr::select(ID, Description) %>%
  dplyr::distinct(.)

# bothP
perSample_GOBP_enrich_df_signif_collapse_bothP <- dplyr::bind_rows(perSample_GOBP_enrich_df_signif_bothP) %>%
  dplyr::select(ID, Description) %>%
  dplyr::distinct(.)

perSample_GOBP_enrich_signif_intersect_df <- as.data.frame(perSample_GOBP_enrich_signif_intersect) %>%
  dplyr::mutate(overlaps = rowSums(.)) %>%
  dplyr::mutate(NFATc1_mut_Only = ifelse(overlaps == 1 & NFATc1_mut == 1, TRUE, FALSE)) %>%
  tibble::rownames_to_column(var="ID") %>%
  dplyr::left_join(., perSample_GOBP_enrich_df_signif_collapse, by = "ID") %>%
  dplyr::arrange(desc(NFATc1_mut_Only))

# bothP
perSample_GOBP_enrich_signif_intersect_df_bothP <- as.data.frame(perSample_GOBP_enrich_signif_intersect_bothP) %>%
  dplyr::mutate(overlaps = rowSums(.)) %>%
  dplyr::mutate(NFATc1_mut_bothP_Only = ifelse(overlaps == 1 & NFATc1_mut_bothP == 1, TRUE, FALSE)) %>%
  tibble::rownames_to_column(var="ID") %>%
  dplyr::left_join(., perSample_GOBP_enrich_df_signif_collapse_bothP, by = "ID") %>%
  dplyr::arrange(desc(NFATc1_mut_bothP_Only))

perSample_GOBP_enrich_signif_intersect_df_NFATc1_mut <- perSample_GOBP_enrich_signif_intersect_df %>%
  dplyr::filter(overlaps == 1 & NFATc1_mut == 1) %>%
  dplyr::left_join(., perSample_GOBP_enrich_df_signif$NFATc1_mut, by = c("ID", "Description"))

perSample_GOBP_enrich_signif_intersect_df_WT <- perSample_GOBP_enrich_signif_intersect_df %>%
  dplyr::filter(overlaps == 1 & WT == 1) %>%
  dplyr::left_join(., perSample_GOBP_enrich_df_signif$WT, by = c("ID", "Description"))

# bothP
perSample_GOBP_enrich_signif_intersect_df_NFATc1_mut_bothP <- perSample_GOBP_enrich_signif_intersect_df_bothP %>%
  dplyr::filter(overlaps == 1 & NFATc1_mut_bothP == 1) #%>%
  dplyr::left_join(., perSample_GOBP_enrich_df_signif_bothP$NFATc1_mut_bothP, by = c("ID", "Description"))

perSample_GOBP_enrich_signif_intersect_df_WT_bothP <- perSample_GOBP_enrich_signif_intersect_df_bothP %>%
  dplyr::filter(overlaps == 1 & WT == 1) %>%
  dplyr::left_join(., perSample_GOBP_enrich_df_signif_bothP$WT, by = c("ID", "Description"))

# perSample_GOBP_enrich_signif_intersect_df_C6 <- perSample_GOBP_enrich_signif_intersect_df %>%
#   dplyr::filter(overlaps == 1 & C6 == 1) %>%
#   dplyr::left_join(., perSample_GOBP_enrich_df_signif$C6, by = "ID")
  

# Plotting top 25 significant GOBP in WT
# perSample_GOBP_enrich_signif_intersect_df_WT

#perSample_GOBP_enrich_signif_intersect_df_WT_ntop <- as.data.frame(perSample_GOBP_enrich_list$WT) #%>%
ntop_signif_gsea = 20
padj_cutoff_ORA = 0.05
perSample_GOBP_enrich_signif_intersect_df_WT_ntop <- perSample_GOBP_enrich_signif_intersect_df_WT %>%
    dplyr::filter(p.adjust < padj_cutoff_ORA) %>%  # significantly enriched
    dplyr::arrange(p.adjust) %>%
    dplyr::slice_head(n = ntop_signif_gsea) %>%
    tidyr::separate(., GeneRatio, sep="/", into = c("GeneRatio_num", "GeneRatio_denom")) %>%
    dplyr::mutate(GeneRatio = as.numeric(GeneRatio_num)/as.numeric(GeneRatio_denom),
                    log10_padj = -log10(p.adjust)) %>%
  dplyr::mutate(Description = stringr::str_wrap(Description, width = 50)) # wrapping long text df$newx = str_wrap(df$x, width = 10)
  
#enrichplot::dotplot(perSample_GOBP_enrich_list$WT, showCategory=25) + ggtitle("dotplot for ORA")
#renv::install("forcats")
perSample_GOBP_enrich_signif_intersect_df_WT_ntop_plot <- ggplot(data = perSample_GOBP_enrich_signif_intersect_df_WT_ntop, 
                                                                 mapping=aes(x = log10_padj, 
                                                                             y = forcats::fct_reorder(Description, log10_padj),
                                                                             color = GeneRatio,
                                                                             size = Count)) +
  geom_point() + 
  theme_bw() +
  scale_colour_gradient(low = "yellow", high = "red", na.value = NA) +
  ggtitle(paste0("Top 20 significantly (p.adj < ", padj_cutoff_ORA,") enriched GOBP terms in WT unique peaks.")) +
  xlab("-log10(padj)") +
  theme(text = element_text(size=12),
        plot.title = element_text(hjust = 0.5), 
        axis.title.y =  element_blank(), 
        axis.text.y = element_text(face = "bold")) 

ggsave(perSample_GOBP_enrich_signif_intersect_df_WT_ntop_plot, filename = paste0(results_dir, "/perSample_GOBP_enrich_signif_WTonly_peaks_ntop20.png"), height = 20, width = 25, units = "cm")
ggsave(perSample_GOBP_enrich_signif_intersect_df_WT_ntop_plot, filename = paste0(results_dir, "/perSample_GOBP_enrich_signif_WTonly_peaks_ntop20.pdf"), height = 20, width = 25, units = "cm")

# adding overlaps to significant results and saving
perSample_GOBP_enrich_df_signif_overlaps <- list(signifGO_overlaps = perSample_GOBP_enrich_signif_intersect_df,
                                                 NFATc1_mut = perSample_GOBP_enrich_signif_intersect_df_NFATc1_mut,
                                                 WT = perSample_GOBP_enrich_signif_intersect_df_WT)

# bothP
perSample_GOBP_enrich_df_signif_overlaps_bothP <- list(signifGO_overlaps = perSample_GOBP_enrich_signif_intersect_df_bothP,
                                                 NFATc1_mut = perSample_GOBP_enrich_signif_intersect_df_NFATc1_mut_bothP,
                                                 WT = perSample_GOBP_enrich_signif_intersect_df_WT_bothP)

message("Saving perSample enrichment results")
save(unique_genes_toEnrich, perSample_GOBP_enrich_list, perSample_GOBP_enrich_df_signif, perSample_GOBP_enrich_signif_intersect_df, file = paste0(results_dir, "/perSample_enrichment.RData"))

save(unique_genes_toEnrich_bothP, perSample_GOBP_enrich_list_bothP, perSample_GOBP_enrich_df_signif_bothP, perSample_GOBP_enrich_signif_intersect_df_bothP, file = paste0(results_dir, "/perSample_enrichment_bothP.RData"))

openxlsx::write.xlsx(x = perSample_GOBP_enrich_df, 
                     file = paste0(results_dir, "/perSample_GOBP_enrich_all.xlsx"),
                     asTable = TRUE,
                     overwrite = TRUE) 

openxlsx::write.xlsx(x = perSample_GOBP_enrich_df_signif_overlaps, 
                     file = paste0(results_dir, "/perSample_GOBP_enrich_signif_overlaps.xlsx"),
                     asTable = TRUE,
                     overwrite = TRUE) 


# bothP
openxlsx::write.xlsx(x = perSample_GOBP_enrich_df_bothP, 
                     file = paste0(results_dir, "/perSample_bothP_GOBP_enrich_all.xlsx"),
                     asTable = TRUE,
                     overwrite = TRUE) 

openxlsx::write.xlsx(x = perSample_GOBP_enrich_df_signif_overlaps_bothP, 
                     file = paste0(results_dir, "/perSample_bothP_GOBP_enrich_signif_overlaps.xlsx"),
                     asTable = TRUE,
                     overwrite = TRUE) 
```

```{r differential accesibility analysis prep, include=TRUE, eval=TRUE, cache=TRUE}
# Defining a set of non-reduntant peaks present in at least 2 samples, count reads in those peaks and use for differential accessibility analysis

n_occurences <- length(sample_names) -1 # peaks need to be present int 3 and more samples
n_occurences_max <- length(sample_names) 
# overlap?
#TODO:
# [ ] - different ways how to find consensus - document how it is done here?! - reduce and overlaps!

if (!file.exists(paste0(results_dir, "/atacseq_metadata_counts.RData"))){
  # Counting for differential ATACseq ----
  nrToCount <- allPeaksSet_filt_nR_annot_gr # allPeaksSet_filt_nR
  occurrences <- rowSums(as.data.frame(elementMetadata(nrToCount))[1:length(sample_names)])
  
  # in two and more samples!
  length(nrToCount)
  consensusToCount <- nrToCount[occurrences >= n_occurences, ]  
  names(consensusToCount) <- consensusToCount$unique_geneID
  length(consensusToCount)
  
  consensusToSizeFactors <- nrToCount[occurrences == n_occurences_max, ]  
  names(consensusToSizeFactors) <- consensusToSizeFactors$unique_geneID
  length(consensusToSizeFactors)
  
  # looking at annotated regions 
  consensusToCount_PID1185 <- consensusToCount[consensusToCount$PID1185 == TRUE,]
  consensusToCount_PID1180 <- consensusToCount[consensusToCount$PID1180 == TRUE,]
  consensusToCount_ND_RJH <- consensusToCount[consensusToCount$ND_RJH == TRUE,]
  consensusToCount_ND_MK <- consensusToCount[consensusToCount$ND_MK == TRUE,]

  consensusToCount_test_grlist <- GenomicRanges::GRangesList(PID1185=consensusToCount_PID1185, 
                                                             PID1180=consensusToCount_PID1180, 
                                                             ND_RJH=consensusToCount_ND_RJH,
                                                             ND_MK=consensusToCount_ND_MK)
  consensus_test_annot <- lapply(consensusToCount_test_grlist, ChIPseeker::annotatePeak, 
                                 TxDb=txdb,
                                 tssRegion=c(-3000, 3000),
                                 level = "gene",  
                                 annoDb="org.Hs.eg.db", 
                                 verbose=FALSE)
  
  consensus_test_annot_bar_plot <- ChIPseeker::plotAnnoBar(consensus_test_annot)
  consensus_test_annot_distTSS_plot <- ChIPseeker::plotDistToTSS(consensus_test_annot,
              title="Distribution of ATAC-seq peaks \nrelative to TSS")
  ggsave(consensus_test_annot_bar_plot, filename = file.path(results_dir, "/consensus_annot_bar_plot.png"))
  ggsave(consensus_test_annot_bar_plot, filename = file.path(results_dir, "/consensus_annot_bar_plot.pdf"))
  
  # counting reads ----
  # to control memory we specify the number of reads to be held in memory at one time using the yield() parameter.
  bamFL <- Rsamtools::BamFileList(project_metadata$bam_files_path, yieldSize = 5e+06)
  
  #TODO:
  # [ ] - import:: Rsamtools functions below
  counts_toSizeFactors <- GenomicAlignments::summarizeOverlaps(features = consensusToSizeFactors, 
                                                               reads = bamFL, 
                                                               singleEnd = TRUE,
                                                               ignore.strand = TRUE,
                                                               fragments = FALSE,
                                                               param = Rsamtools::ScanBamParam(flag = Rsamtools::scanBamFlag(isDuplicate = FALSE)))
  counts_consensus <- GenomicAlignments::summarizeOverlaps(features = consensusToCount, 
                                                           reads = bamFL, 
                                                           singleEnd = TRUE,
                                                           ignore.strand = TRUE,
                                                           fragments = FALSE,
                                                           param = Rsamtools::ScanBamParam(flag = Rsamtools::scanBamFlag(isDuplicate = FALSE)))
  
  colnames(counts_toSizeFactors) <- sample_names
  colnames(counts_consensus) <- sample_names
  
  
  # exporting consensus counts ---
  #AAA <- as.data.frame(rowData(counts_consensus))
  #BBB <- as.data.frame(rowRanges(counts_consensus))
  # identical(AAA, BBB) yes after seqnames, start, end, width and strand is removed
  #AAA <- as.data.frame(counts_consensus@rowRanges)
  
  counts_consensus_ranges_df <- as.data.frame(rowRanges(counts_consensus)) %>%
    dplyr::rename(PID1185_peak_detected = PID1185,
                  PID1180_peak_detected = PID1180,
                  ND_RJH_peak_detected = ND_RJH,
                  ND_MK_peak_detected = ND_MK)
  
  # save this in a tsv - consensus_peaks_counts.tsv and generate md5sums!
  counts_consensus_counts_df <- as.data.frame(assay(counts_consensus)) %>%  # dim 21627 3
    tibble::rownames_to_column(var="unique_geneID") %>%
    dplyr::rename(PID1185_counts = PID1185,
                  PID1180_counts = PID1180,
                  ND_RJH_counts = ND_RJH,
                  ND_MK_counts = ND_MK) %>%
    dplyr::left_join(., counts_consensus_ranges_df, by = "unique_geneID") %>%
    # renaming samples:
    # ND_01 ->	C5
    # ND_02	-> C6
    # Patient ->	P
    # !!! Incorporate in the code above!
    dplyr::rename(PID1185_counts = PID1185_counts,
                  PID1180_counts = PID1180_counts,
                  ND_RJH_counts = ND_RJH_counts,
                  ND_MK_counts = ND_MK_counts,
                  PID1185_peak_detected = PID1185_peak_detected,
                  PID1180_peak_detected = PID1180_peak_detected,
                  ND_RJH_peak_detected = ND_RJH_peak_detected,
                  ND_MK_peak_detected= ND_MK_peak_detected) %>%
    dplyr::select(unique_geneID, seqnames, start, end, width, strand, 
                  PID1185_peak_detected, PID1180_peak_detected, ND_RJH_peak_detected, ND_MK_peak_detected,
                  PID1185_counts, PID1180_counts, ND_RJH_counts, ND_MK_counts,
                  dplyr::everything())

  readr::write_tsv(x = counts_consensus_counts_df,
                   file = file.path(results_dir, "atacseq_consensus_peaks_counts.tsv"))
  
  
  #rowRanges(myCounts)
  
  #count_data <- assay(myCounts)
  all(colnames(counts_toSizeFactors) == rownames(project_metadata))
  all(colnames(counts_consensus) == rownames(project_metadata))
  
  message("Saving metadata and counts")
  save(project_metadata, counts_consensus, counts_toSizeFactors, nrToCount, blacklist, file = paste0(results_dir, "/atacseq_metadata_counts.RData"))
} else {
  message("Loading metadata and counts")
  load(file = paste0(results_dir, "/atacseq_metadata_counts.RData"))
}


```

```{r differential accesibility analysis run, include=TRUE, eval=TRUE, cache=TRUE}

# DESeq2 analysis ----
#TODO:
# [ ] - what are the control genes defined below?!

if (!file.exists(paste0(results_dir, "/dds_objects.RData"))){
  # pre-calculating size factors
  project_metadata_upd <- project_metadata %>%
    dplyr::mutate(genotype = factor(genotype, levels = c("WT", "NFATc1_mut")))
  atacDDS_toSizeFactors <- DESeq2::DESeqDataSetFromMatrix(countData = assay(counts_toSizeFactors), 
                                                          colData = project_metadata_upd, 
                                                          design = ~genotype, 
                                                          rowRanges = rowRanges(counts_toSizeFactors))  # rowRanges(myCounts) == 
  
  atacDDS_withoutControls <- DESeq2::DESeqDataSetFromMatrix(countData = assay(counts_consensus), 
                                                            colData = project_metadata_upd, 
                                                            design = ~genotype, 
                                                            rowRanges = rowRanges(counts_consensus))  # rowRanges(myCounts) == consensusToCount
  
  atacDDS <- DESeq2::DESeqDataSetFromMatrix(countData = assay(counts_consensus), 
                                            colData = project_metadata_upd, 
                                            design = ~genotype, 
                                            rowRanges = rowRanges(counts_consensus))  # rowRanges(myCounts) == consensusToCount
  
  atacDDS_toSizeFactors <- DESeq2::estimateSizeFactors(atacDDS_toSizeFactors)
  atacDDS_withoutControls <- DESeq2::estimateSizeFactors(atacDDS_withoutControls)
  
  atacDDS_toSizeFactors_sf <- DESeq2::sizeFactors(atacDDS_toSizeFactors)
  atacDDS_withoutControls_sf <- DESeq2::sizeFactors(atacDDS_withoutControls)
  
  # frozen geomMeans? (but would need all of the gens)
  # rather use controlGenes
  control_genes <- rownames(atacDDS_withoutControls) %in% rownames(atacDDS_toSizeFactors)
  table(control_genes)
  atacDDS <- DESeq2::estimateSizeFactors(atacDDS, controlGenes = control_genes)
  atacDDS_sf <- DESeq2::sizeFactors(atacDDS)
  
  # comparing size factors
  sf_comparison_list <- list(controls_sf = atacDDS_toSizeFactors_sf,
                             consensus_noControls = atacDDS_withoutControls_sf,
                             consensus_controlsCorrected = atacDDS_sf)
  sf_comparison <- dplyr::bind_rows(sf_comparison_list, .id="sf_origin")
  
  # different length :D
  # geoMeans <- exp(rowMeans(log(counts(atacDDS_toSizeFactors))))
  # dds_test <- estimateSizeFactors(atacDDS,geoMeans=geoMeans)
  # sizeFactors(dds)
  
  # plot before and after normalization and differences in using all overlapping size factors?!
  atacDDS_withoutControls_rawCounts <- DESeq2::counts(object = atacDDS_withoutControls, normalized = FALSE)
  atacDDS_withoutControls_normCounts <- DESeq2::counts(object = atacDDS_withoutControls, normalized = TRUE)
  atacDDS_rawCounts <- DESeq2::counts(object = atacDDS, normalized = FALSE)
  atacDDS_normCounts <- DESeq2::counts(object = atacDDS, normalized = TRUE)
  
  atacDDS_withoutControls_rawCounts_df <- as.data.frame(atacDDS_withoutControls_rawCounts) %>%
    tibble::rownames_to_column(var = "unique_geneID") %>%
    tidyr::pivot_longer(., -unique_geneID, names_to = "sample_name", values_to = "counts") %>%
    dplyr::mutate(count_type = "raw_counts",
                  sf_origin = "consensus_noControls")
  
  atacDDS_withoutControls_normCounts_df <- as.data.frame(atacDDS_withoutControls_normCounts) %>%
    tibble::rownames_to_column(var = "unique_geneID") %>%
    tidyr::pivot_longer(., -unique_geneID, names_to = "sample_name", values_to = "counts") %>%
    dplyr::mutate(count_type = "norm_counts",
                  sf_origin = "consensus_noControls")
  
  atacDDS_rawCounts_df <- as.data.frame(atacDDS_rawCounts) %>%
    tibble::rownames_to_column(var = "unique_geneID") %>%
    tidyr::pivot_longer(., -unique_geneID, names_to = "sample_name", values_to = "counts") %>%
    dplyr::mutate(count_type = "raw_counts",
                  sf_origin = "consensus_controlsCorrected")
  
  atacDDS_normCounts_df <- as.data.frame(atacDDS_normCounts) %>%
    tibble::rownames_to_column(var = "unique_geneID") %>%
    tidyr::pivot_longer(., -unique_geneID, names_to = "sample_name", values_to = "counts") %>%
    dplyr::mutate(count_type = "norm_counts",
                  sf_origin = "consensus_controlsCorrected")
  
  sf_comparison_counts_df <- dplyr::bind_rows(atacDDS_withoutControls_rawCounts_df,
                                              atacDDS_withoutControls_normCounts_df,
                                              atacDDS_rawCounts_df,
                                              atacDDS_normCounts_df) %>%
    dplyr::mutate(count_type = factor(count_type, levels = c("raw_counts", "norm_counts")),
                  sf_origin = factor(sf_origin, levels = c("consensus_noControls", "consensus_controlsCorrected")))
  
  sf_comparison_counts_box_plot <- ggplot(data = sf_comparison_counts_df, aes(x=count_type, y=log2(counts + 1), fill=sample_name)) + 
    geom_boxplot() + 
    facet_wrap(~sf_origin) +
    theme_bw()
  
  sf_comparison_counts_dens_plot <- ggplot(data = sf_comparison_counts_df, aes(x=log2(counts + 1), fill=sample_name)) + 
    geom_density(alpha = 0.6) + 
    facet_wrap(~count_type + sf_origin) +
    theme_bw()
  
  ggsave(sf_comparison_counts_box_plot, filename = file.path(results_dir, "/sf_comparison_counts_box_plot.png"))
  ggsave(sf_comparison_counts_dens_plot, filename = file.path(results_dir, "/sf_comparison_counts_dens_plot.png"))
  
  # betaPrior=FALSE is now by default and should be followed by lfcShrink
  # however, setting here betaPrior=T; also consistent with similar setting as Florian: dds <- DESeq(dds, test="Wald", fitType="local", betaPrior=T)
  
  # further compare
  #atacDDS <- DESeq2::DESeq(atacDDS, test="Wald", fitType="local", betaPrior=T)
  
  # here they also run withut betaPrior? check version!
  # https://github.com/Zhang-lab/ATACseq_benchmarking/blob/master/simulated_ATAC_tests_DESeq2.R
  # https://support.bioconductor.org/p/95436/
  # https://support.bioconductor.org/p/68772/
  # https://bioinformatics.stackexchange.com/questions/10837/loss-and-gain-of-region-in-atac-seq
  #https://support.bioconductor.org/p/103348/
  # also setting betaPrior=TRUE since no shrinking is done aferwards; https://support.bioconductor.org/p/118257/
  # make sure it uses pre-existing size factors!
  atacDDS_withoutControls <- DESeq2::DESeq(atacDDS_withoutControls, test="Wald", fitType="local", betaPrior=TRUE)
  atacDDS <- DESeq2::DESeq(atacDDS, test="Wald", fitType="local", betaPrior=TRUE)  # set local as in Diffbind; see manual and 
  
  # what dispersion is expected for ATAC-seq?]
  # par(mfrow=c(1,2)) 
  # DESeq2::plotDispEsts(atacDDS_withoutControls)
  # DESeq2::plotDispEsts(atacDDS)
  # dev.off()
  
  # error in fit for rlog! if fitType="local" not specified 
  # -- note: fitType='parametric', but the dispersion trend was not well captured by the
  #  function: y = a/x + b, and a local regression fit was automatically substituted.
  #  specify fitType='local' or 'mean' to avoid this message next time.
  atac_rlog <- DESeq2::rlog(atacDDS, fitType="local")
  atac_rlog_ranges <- rowRanges(atac_rlog)
  atac_rlog_counts <- assay(atac_rlog)  # use for heatmaps
  atac_pca <- DESeq2::plotPCA(atac_rlog, intgroup = "genotype", ntop = nrow(atac_rlog)) + theme_bw()
  
  message("Saving dds objects")
  save(atacDDS, atacDDS_withoutControls, atacDDS_toSizeFactors, file = paste0(results_dir, "/dds_objects.RData"))
} else {
  
  message("Loading dds objects")
  load(file = paste0(results_dir, "/dds_objects.RData"))
  
  atac_rlog <- DESeq2::rlog(atacDDS, fitType="local")
  atac_rlog_ranges <- rowRanges(atac_rlog)
  atac_rlog_counts <- assay(atac_rlog)
}


# With the new DESeq2 object we can now test for any differences in ATACseq signal between groups.

#DESeq2::resultsNames(atacDDS)

# extracting differential accessibility results ----
# contrasts
# patient_vs_NDs c("sample_group", "Patient", "ND")
#project_metadata_upd
# DESeq2::resultsNames(atacDDS)
de_comparisons <- list(NFATc1_mut_vs_WT = c("genotype", "NFATc1_mut", "WT"))

de_results <- purrr::map(.x = names(de_comparisons), .f = function(comparison_name){
  # extract DE results
  #   run annotation
  # extract significant - UP, DOWN
  #   run annotation
  #   run GO enrichment
  
  message(paste0("Processing comparison: ", comparison_name))
  
  # extract DE results
  #   run annotation
  temp_contrast <- de_comparisons[[comparison_name]]
  de_results_gr <- DESeq2::results(atacDDS, contrast = temp_contrast, format = "GRanges")
  de_results_gr <- de_results_gr[order(de_results_gr$pvalue)]  # order based on p-value
  
  return(list(de_results_gr = de_results_gr))
  
}) %>% setNames(names(de_comparisons))


# NFATc1_mut_vs_WT ----
# annotating DESeq2 results
# selecting significant results

NFATc1_mut_vs_WT_results_dir <- file.path(results_dir, "NFATc1_mut_vs_WT_results")
dir.create(NFATc1_mut_vs_WT_results_dir, showWarnings = TRUE)
NFATc1_mut_vs_WT_results_motif_dir <- file.path(NFATc1_mut_vs_WT_results_dir, "motif_analysis")
dir.create(NFATc1_mut_vs_WT_results_motif_dir, showWarnings = TRUE)

# all peaks, not just consensus; but not using normalized counts
NFATc1_mut_vs_WT_results_motif_allPeaks_dir <- file.path(NFATc1_mut_vs_WT_results_dir, "motif_analysis_allPeaks")
dir.create(NFATc1_mut_vs_WT_results_motif_allPeaks_dir, showWarnings = TRUE)

# annotating DESeq2 results ----
de_results_gr <- de_results$NFATc1_mut_vs_WT$de_results_gr
de_results_df <- as.data.frame(de_results_gr) %>%
  tibble::rownames_to_column(var="unique_geneID") %>%
  #dplyr::mutate(geneId = stringr::str_split_fixed(unique_geneID, pattern = "_", 4)[,1]) %>%
  dplyr::left_join(., dplyr::select(allPeaksSet_filt_nR_annot_df, -seqnames, -start, -end, -width, -strand), by = "unique_geneID")  # annotating

#TODO:
# [ ] - save intermediate results!!!!

# selecting signaficance results ----

# padj significance
hist(de_results_df$pvalue)
hist(de_results_df$padj)
#FIXME:
# [ ] - using pvalue rather padj!!!!
# de_results_df_filt_signifPadj <- de_results_df %>%
#   dplyr::filter(!is.na(padj) & padj < padj_cutoff)
# de_results_df_filt_signifPadj <- de_results_df %>%
#   dplyr::filter(!is.na(padj) & pvalue < padj_cutoff)
# de_results_df_filt_signifPval <- de_results_df %>%
#   dplyr::filter(!is.na(padj) & pvalue < padj_cutoff)
# quantile(de_results_df_filt_signifPadj$log2FoldChange)  # these are all down!
  
# de_results_df_filt_signifPadj_UP_DOWN <- de_results_df_filt_signifPval %>%
#   dplyr::mutate(FC_direction = ifelse(log2FoldChange < 0, "DOWN", "UP"))
# table(de_results_df_filt_signifPadj_UP_DOWN$FC_direction)

de_results_df_filt_signifLog2Promoters_df <- de_results_df %>%
  dplyr::filter((abs(log2FoldChange) > log2FC_cutoff) & (abs(distanceToTSS) <= promoter_dist))

de_results_df_filt_signifLog2Promoters_df_UP_DOWN <- de_results_df_filt_signifLog2Promoters_df %>%
  dplyr::mutate(FC_direction = ifelse(log2FoldChange < 0, "DOWN", "UP"))
table(de_results_df_filt_signifLog2Promoters_df_UP_DOWN$FC_direction)

# renaming duplicates
de_results_df_filt_signifLog2Promoters_df_dupl <- de_results_df_filt_signifLog2Promoters_df[duplicated(de_results_df_filt_signifLog2Promoters_df$SYMBOL) | duplicated(de_results_df_filt_signifLog2Promoters_df$SYMBOL, fromLast=TRUE),]

de_results_df_filt_signifLog2Promoters_df_duplFix <- de_results_df_filt_signifLog2Promoters_df %>%
  dplyr::mutate(SYMBOL = ifelse(SYMBOL %in% de_results_df_filt_signifLog2Promoters_df_dupl$SYMBOL, paste(SYMBOL, seqnames, start, end, sep = "_"), SYMBOL))

# for heatmap ----
# mostly down
# de_results_df_filt_signifPadj_forHeatmap <- de_results_df_filt_signifPadj %>%
#   dplyr::select(unique_geneID, SYMBOL)
# check for duplicates in gene symbols
#sum(duplicated(de_results_df_filt_signifPadj_forHeatmap$SYMBOL))
  
# atac_rlog_counts_forHeatmap_signifPadj <- as.data.frame(atac_rlog_counts) %>%
#   tibble::rownames_to_column(var="unique_geneID") %>%
#   dplyr::right_join(., de_results_df_filt_signifPadj_forHeatmap, by = "unique_geneID") %>%
#   dplyr::select(-unique_geneID) %>%
#   tibble::column_to_rownames("SYMBOL")
#   
annotCol_forHeatmap <- project_metadata %>%
  dplyr::select(sample_name, condition) %>%
  tibble::column_to_rownames(var="sample_name")

annotRow_forHeatmap <- de_results_df_filt_signifPadj %>%
  dplyr::select(SYMBOL, annotation) %>%
  dplyr::mutate(annotation = gsub(pattern = "(Exon|Intron)( \\(.+)", replacement = "\\1", annotation)) %>%
  tibble::column_to_rownames(var="SYMBOL")

pheatmap::pheatmap(atac_rlog_counts_forHeatmap_signifPadj, 
                   scale = "row", 
                   cluster_cols = FALSE,
                   annotation_col = annotCol_forHeatmap,
                   annotation_row = annotRow_forHeatmap)

pheatmap::pheatmap(atac_rlog_counts_forHeatmap_signifPadj, 
                   scale = "row", 
                   cluster_cols = FALSE,
                   annotation_col = annotCol_forHeatmap,
                   annotation_row = annotRow_forHeatmap,
                  filename = file.path(NFATc1_mut_vs_WT_results_dir, "de_results_heatmap_signifPadj.pdf"))

pheatmap::pheatmap(atac_rlog_counts_forHeatmap_signifPadj, 
                   scale = "row", 
                   cluster_cols = FALSE,
                   annotation_col = annotCol_forHeatmap,
                   annotation_row = annotRow_forHeatmap,
                  filename = file.path(NFATc1_mut_vs_WT_results_dir, "de_results_heatmap_signifPadj.png"))

# for logFC significance
de_results_df_filt_signifLog2Promoters_forHeatmap <- de_results_df_filt_signifLog2Promoters_df_duplFix %>%
  dplyr::select(unique_geneID, SYMBOL)
# check for duplicates in gene symbols
sum(duplicated(de_results_df_filt_signifLog2Promoters_forHeatmap$SYMBOL))

# re-run with blind=FALSE
atac_rlog <- DESeq2::rlog(atacDDS, fitType="local", blind=FALSE)
atac_rlog_ranges <- rowRanges(atac_rlog)
atac_rlog_counts <- assay(atac_rlog)
  
atac_rlog_counts_forHeatmap_signifLog2Promoters <- as.data.frame(atac_rlog_counts) %>%
  tibble::rownames_to_column(var="unique_geneID") %>%
  dplyr::right_join(., de_results_df_filt_signifLog2Promoters_forHeatmap, by = "unique_geneID") %>%
  dplyr::select(-unique_geneID) %>%
  tibble::column_to_rownames("SYMBOL")
  
  
annotCol_forHeatmap <- project_metadata_upd %>%
  dplyr::select(sample_name, genotype) %>%
  tibble::column_to_rownames(var="sample_name")

annotRow_forHeatmap <- de_results_df_filt_signifLog2Promoters_df_duplFix %>%
  dplyr::select(SYMBOL, annotation) %>%
  dplyr::mutate(annotation = gsub(pattern = "(Exon|Intron)( \\(.+)", replacement = "\\1", annotation)) %>%
  tibble::column_to_rownames(var="SYMBOL")

atac_rlog_counts_forHeatmap_signifLog2Promoters_reord <- atac_rlog_counts_forHeatmap_signifLog2Promoters[c("ND_RJH","ND_MK", "PID1185", "PID1180")]
pheatmap::pheatmap(atac_rlog_counts_forHeatmap_signifLog2Promoters_reord, 
                   scale = "row", 
                   cluster_cols = FALSE,
                   annotation_col = annotCol_forHeatmap,
                   annotation_row = annotRow_forHeatmap,
                   fontsize_row = 10)  # fontsize_row = 10

pheatmap::pheatmap(atac_rlog_counts_forHeatmap_signifLog2Promoters_reord, 
                   scale = "row", 
                   cluster_cols = FALSE,
                   annotation_col = annotCol_forHeatmap,
                   annotation_row = annotRow_forHeatmap,
                   fontsize_row = 10,
                   filename = file.path(NFATc1_mut_vs_WT_results_dir, "de_results_heatmap_signifLog2Promoters.pdf"))

pheatmap::pheatmap(atac_rlog_counts_forHeatmap_signifLog2Promoters_reord, 
                   scale = "row", 
                   cluster_cols = FALSE,
                   annotation_col = annotCol_forHeatmap,
                   annotation_row = annotRow_forHeatmap,
                   fontsize_row = 10,
                  filename = file.path(NFATc1_mut_vs_WT_results_dir, "de_results_heatmap_signifLog2Promoters.png"))

# pub quality heatmap ----
# for the downregulated: HDAC7, NT5E, IL7R, CCR2, LDOC1, CRTAM, IL2, ITGAX, TESPA1, TRAF2IP2, STAT5A, TCF7, IFNG, IKZF2, CD44, CAP70, RANGAP1, LY9, NFAT5, CDK6, CD7 and VAV1
# for the upregulated: FOXO6, KLF16, SUB1, IL7, TGFB1, HEXD, NELFCD, CSNK1G2, MAP2K7, PPP1R15A and TCF25
# highlight_genes <- c("HDAC7", "NT5E", "IL7R", "CCR2", "LDOC1", "CRTAM", "IL2", "ITGAX", "TESPA1", "TRAF3IP2", "STAT5A", "TCF7", "IFNG", "IKZF2", "CD44", "ZAP70", "RANGAP1", "LY9", "NFAT5", "CDK6", "CD7", "VAV1", "FOXO6", "KLF16", "SUB1", "IL7", "TGFB1", "HEXD", "NELFCD", "CSNK1G2", "MAP2K7", "PPP1R15A", "TCF25")
# length(highlight_genes) 33 genes!

annotCol_forHeatmap <- project_metadata %>%
  dplyr::select(sample_name, condition) %>%
  dplyr::rename(genotype = condition) %>%
  dplyr::mutate(genotype = if_else(genotype == "IKZF2-WT", "WT", "I325V-Hom")) %>%
  dplyr::mutate(bio_rep = sample_name) %>%
  tibble::column_to_rownames(var="sample_name")

# annotRow_forHeatmap <- de_results_df_filt_signifPadj %>%
#   dplyr::select(SYMBOL, annotation) %>%
#   dplyr::mutate(annotation = gsub(pattern = "(Exon|Intron)( \\(.+)", replacement = "\\1", annotation)) %>%
#   tibble::column_to_rownames(var="SYMBOL")

#scales::show_col(c("#d9d9d9", "#969696", "#fd8d3c"))

annotCol_forHeatmap_colors <- list(genotype = c(  WT="#525252", `I325V-Hom` = "#fc4e2a"),
                                   bio_rep = c(C5 = "#d9d9d9", C6= "#969696", P = "#fd8d3c"))

# - [] large gap between legend and heatmap - due to long gene names!
# remove row labels during plotting
row_labels <- rownames(atac_rlog_counts_forHeatmap_signifLog2Promoters)
#nchar(row_labels)
row_labels <- ifelse(row_labels %in% highlight_genes, row_labels, stringr::str_pad("", width = 25, side = "both")) # 14 spaces - small hack to keep gaps after repel
#nchar(row_labels)
# small hack to add spacing between legend and heatmap after 

signifLog2Promoters_heatmap <-  pheatmap::pheatmap(as.matrix(atac_rlog_counts_forHeatmap_signifLog2Promoters),
                                                   color = colorRampPalette(rev(RColorBrewer::brewer.pal(7, "PiYG")))(100),
                                                   scale = "row",
                                                   #gaps_col = ph_col_annot%>% arrange(state)%$% table(state) %>% cumsum,
                                                   #gaps_row = ph_genes$clust %>% table() %>% cumsum,
                                                   cluster_cols = FALSE,
                                                   cluster_rows = TRUE,
                                                   show_colnames = FALSE,
                                                   show_rownames = TRUE,
                                                   annotation_col = annotCol_forHeatmap,
                                                   #annotation_row = ph_row_annot,
                                                   #annotation_legend = FALSE,
                                                   annotation_colors = annotCol_forHeatmap_colors,  # add manual annotation corresponding to samples
                                                   labels_row = row_labels,
                                                   fontsize_row = 9,
                                                   fontsize = 9,
                                                   #cellheight = 9,
                                                   main = "Significantly differentially accessible regions")

signifLog2Promoters_heatmap_upd <- add_flag(signifLog2Promoters_heatmap,
                                            kept.labels = highlight_genes,
                                            repel.degree = 0.2,
                                            hiden.labels.character = "") # 14 spaces? - small hack to keep gaps after repel
  
pdf(file.path(NFATc1_mut_vs_WT_results_dir, "dar_signifLog2Promoters_heatmap.pdf"))
grid.draw(signifLog2Promoters_heatmap_upd)
dev.off()
  
# also add filtering criteria!
message("Saving DE results")

load(file.path(NFATc1_mut_vs_WT_results_dir, "de_results_all.RData"))

# de_filters <- data.frame(de_results_signifPadj = paste0("!is.na(padj) & padj < ", padj_cutoff),
#                          de_results_signifLog2Promoters = paste0("(abs(log2FoldChange) > ", log2FC_cutoff,") & (abs(distanceToTSS) <= ", promoter_dist,"); distanceToTSS - window of ", promoter_dist," around TSS"))

# de_results_df_toSave <- list(de_results = de_results_df,
#                              de_results_signifPadj = de_results_df_filt_signifPadj,
#                              de_results_signifLog2Promoters = de_results_df_filt_signifLog2Promoters_df,
#                              applied_filters = de_filters)

de_filters <- data.frame(de_results_signifLog2Promoters = paste0("(abs(log2FoldChange) > ", log2FC_cutoff,") & (abs(distanceToTSS) <= ", promoter_dist,"); distanceToTSS - window of ", promoter_dist," around TSS"))

de_results_df_toSave <- list(de_results = de_results_df,
                             de_results_signifLog2Promoters = de_results_df_filt_signifLog2Promoters_df,
                             applied_filters = de_filters)
openxlsx::write.xlsx(x = de_results_df_toSave, 
                     file = paste0(NFATc1_mut_vs_WT_results_dir, "/da_results_NFATc1_mut_vs_WT_results.xlsx"),
                     asTable = TRUE,
                     overwrite = TRUE) 

#   run annotation
# extract significant - UP, DOWN
#   run annotation
#   run GO enrichment
# significant based on log2FC

enrich_GOterms <- c(GOBP="BP", GOMF="MF", GOCC="CC")

# de_results_df_filt_signifPadj_UP <- de_results_df_filt_signifPadj %>%
#   dplyr::filter(log2FoldChange > 0)
# de_results_df_filt_signifPadj_DOWN <- de_results_df_filt_signifPadj %>%
#   dplyr::filter(log2FoldChange < 0)
#   
# message("de_results_signif_annot_df all: ", nrow(de_results_df_filt_signifPadj))  
# message("de_results_signif_annot_df_UP genes: ", nrow(de_results_df_filt_signifPadj_UP))
# message("de_results_signif_annot_df_DOWN genes: ", nrow(de_results_df_filt_signifPadj_DOWN))
# 
# signifPadj_genesToEnrich <- list(all=unique(de_results_df_filt_signifPadj$geneId),
#                                           up=unique(de_results_df_filt_signifPadj_UP$geneId),
#                                           down=unique(de_results_df_filt_signifPadj_DOWN$geneId))
# 
# 
# if (!file.exists(file.path(NFATc1_mut_vs_WT_results_dir, "de_results_goEnrich.RData"))) {
#   
#   signifPadj_goEnrichment <- purrr::map(.x = names(signifPadj_genesToEnrich), .f = function(gene_set){
#     message("Enriching genes ", gene_set)
#     
#     goEnrichment <- purrr::map(.x = names(enrich_GOterms), .f = function(GO_name){
#       message("Enriching for ", GO_name)
#       
#       clusterProfiler::enrichGO(gene = signifPadj_genesToEnrich[[gene_set]],
#                                 universe      = gene_universe,
#                                 OrgDb         = org.Hs.eg.db,
#                                 ont           = enrich_GOterms[[GO_name]],  # MF, CC, BP, ALL
#                                 pAdjustMethod = "BH", 
#                                 pvalueCutoff  = 1,
#                                 qvalueCutoff  = 1,
#                                 readable      = TRUE)
#       
#     }) %>% setNames(names(enrich_GOterms))
#     
#     return(goEnrichment)
#     
#   }) %>% setNames(names(signifPadj_genesToEnrich))
#   
#   
#   signifPadj_ALL_goEnrichment_df <- purrr::map(.x = signifPadj_goEnrichment$all, .f = as.data.frame) %>% setNames(names(signifPadj_goEnrichment$all))
#   signifPadj_UP_goEnrichment_df <- purrr::map(.x = signifPadj_goEnrichment$up, .f = as.data.frame) %>% setNames(names(signifPadj_goEnrichment$up))
#   signifPadj_DOWN_goEnrichment_df <- purrr::map(.x = signifPadj_goEnrichment$down, .f = as.data.frame) %>% setNames(names(signifPadj_goEnrichment$down))
#   
#   openxlsx::write.xlsx(x = signifPadj_ALL_goEnrichment_df, 
#                        file = paste0(NFATc1_mut_vs_WT_results_dir, "/de_results_patient_vs_NDs_signifPadj_ALL_GOenrich.xlsx"),
#                        asTable = TRUE,
#                        overwrite = TRUE) 
#   
#   openxlsx::write.xlsx(x = signifPadj_DOWN_goEnrichment_df, 
#                        file = paste0(NFATc1_mut_vs_WT_results_dir, "/de_results_patient_vs_NDs_signifPadj_DOWN_GOenrich.xlsx"),
#                        asTable = TRUE,
#                        overwrite = TRUE) 
  
  # on log2FC ----
  de_results_df_filt_signifLog2Promoters_UP <- de_results_df_filt_signifLog2Promoters_df %>%
    dplyr::filter(log2FoldChange > 0)
  de_results_df_filt_signifLog2Promoters_DOWN <- de_results_df_filt_signifLog2Promoters_df %>%
    dplyr::filter(log2FoldChange < 0)
  
  message("de_results_signif_annot_df all: ", nrow(de_results_df_filt_signifLog2Promoters_df))  
  message("de_results_signif_annot_df_UP genes: ", nrow(de_results_df_filt_signifLog2Promoters_UP))
  message("de_results_signif_annot_df_DOWN genes: ", nrow(de_results_df_filt_signifLog2Promoters_DOWN))
  
  signifLog2Promoters_genesToEnrich <- list(all=unique(de_results_df_filt_signifLog2Promoters_df$geneId),
                                            up=unique(de_results_df_filt_signifLog2Promoters_UP$geneId),
                                            down=unique(de_results_df_filt_signifLog2Promoters_DOWN$geneId))
  
  signifLog2Promoters_goEnrichment <- purrr::map(.x = names(signifLog2Promoters_genesToEnrich), .f = function(gene_set){
    message("Enriching genes ", gene_set)
    
    goEnrichment <- purrr::map(.x = names(enrich_GOterms), .f = function(GO_name){
      message("Enriching for ", GO_name)
      
      clusterProfiler::enrichGO(gene = signifLog2Promoters_genesToEnrich[[gene_set]],
                                universe      = gene_universe,
                                OrgDb         = org.Hs.eg.db,
                                ont           = enrich_GOterms[[GO_name]],  # MF, CC, BP, ALL
                                pAdjustMethod = "BH", 
                                pvalueCutoff  = 1,
                                qvalueCutoff  = 1,
                                readable      = TRUE)
      
    }) %>% setNames(names(enrich_GOterms))
    
    return(goEnrichment)
    
  }) %>% setNames(names(signifLog2Promoters_genesToEnrich))
  
  
  # also add filtering criteria!
  message("Saving GOBP enrichment results")
  
  # save(signifPadj_goEnrichment, signifLog2Promoters_goEnrichment,
  #      file = file.path(NFATc1_mut_vs_WT_results_dir, "de_results_goEnrich.RData"))
  save(signifLog2Promoters_goEnrichment,
       file = file.path(NFATc1_mut_vs_WT_results_dir, "da_results_goEnrich.RData"))
  
} else {
  message("Loading GOBP enrichment results")
  load(file = file.path(NFATc1_mut_vs_WT_results_dir, "de_results_goEnrich.RData"))
  
}

signifLog2Promoters_ALL_goEnrichment_df <- purrr::map(.x = signifLog2Promoters_goEnrichment$all, .f = as.data.frame) %>% setNames(names(signifLog2Promoters_goEnrichment$all))

signifLog2Promoters_UP_goEnrichment_df <- purrr::map(.x = signifLog2Promoters_goEnrichment$up, .f = as.data.frame) %>% setNames(names(signifLog2Promoters_goEnrichment$up))

signifLog2Promoters_DOWN_goEnrichment_df <- purrr::map(.x = signifLog2Promoters_goEnrichment$down, .f = as.data.frame) %>% setNames(names(signifLog2Promoters_goEnrichment$down))

lapply(signifLog2Promoters_ALL_goEnrichment_df, length)
lapply(signifLog2Promoters_UP_goEnrichment_df, length)
lapply(signifLog2Promoters_DOWN_goEnrichment_df, length)

openxlsx::write.xlsx(x = signifLog2Promoters_ALL_goEnrichment_df, 
                     file = paste0(NFATc1_mut_vs_WT_results_dir, "/da_results_NFATc1_mut_vs_WT_signifLog2Promoters_ALL_GOenrich.xlsx"),
                     asTable = TRUE,
                     overwrite = TRUE) 

openxlsx::write.xlsx(x = signifLog2Promoters_UP_goEnrichment_df, 
                     file = paste0(NFATc1_mut_vs_WT_results_dir, "/da_results_NFATc1_mut_vs_WT_signifLog2Promoters_UP_GOenrich.xlsx"),
                     asTable = TRUE,
                     overwrite = TRUE) 

openxlsx::write.xlsx(x = signifLog2Promoters_DOWN_goEnrichment_df, 
                     file = paste0(NFATc1_mut_vs_WT_results_dir, "/da_results_NFATc1_mut_vs_WT_signifLog2Promoters_DOWN_GOenrich.xlsx"),
                     asTable = TRUE,
                     overwrite = TRUE) 

# accessibility heatmap ----
# - [ ] add results from enricheR - they differ a bit; z-scores?

# plotting GO term enrichment
# top 25 and top50
signifLog2Promoters_DOWN_goEnrichment_df <- purrr::map(.x = signifLog2Promoters_goEnrichment$down, .f = as.data.frame) %>% setNames(names(signifLog2Promoters_goEnrichment$down))

signifLog2Promoters_DOWN_goEnrichment_GOBP <- signifLog2Promoters_DOWN_goEnrichment_df$GOBP

# collapse go terms?

ntop_GOTERMS=50
signifLog2Promoters_DOWN_goEnrichment_GOBP_toPlot <- signifLog2Promoters_DOWN_goEnrichment_GOBP %>%
  dplyr::filter(p.adjust < 0.05) %>%
  dplyr::mutate(log10padj = -log(p.adjust, base = 10)) %>%
  dplyr::arrange(desc(log10padj)) 

signifLog2Promoters_DOWN_goEnrichment_GOBP_toPlot$Description <- factor(signifLog2Promoters_DOWN_goEnrichment_GOBP_toPlot$Description, levels = signifLog2Promoters_DOWN_goEnrichment_GOBP_toPlot$Description)

# https://ggplot2.tidyverse.org/reference/geom_bar.html
signifLog2Promoters_DOWN_GOBP_barplot <- ggplot(signifLog2Promoters_DOWN_goEnrichment_GOBP_toPlot[1:ntop_GOTERMS,], aes(x=log10padj, y=fct_rev(Description), fill = log10padj)) + 
  scale_y_discrete(labels = scales::wrap_format(60)) +
  geom_col() +
  theme_bw() +
  xlab("-log10(padj)") + ylab("") +
  labs(fill = "-log10(padj)") + 
  scale_fill_gradient2(low="navy", mid="white", high="red", 
                       midpoint=0)

ggsave(signifLog2Promoters_DOWN_GOBP_barplot, filename = paste0(NFATc1_mut_vs_WT_results_dir, "/signifLog2Promoters_DOWN_GOBP_barplot_ntop",ntop_GOTERMS,".pdf"))
ggsave(signifLog2Promoters_DOWN_GOBP_barplot, filename = paste0(NFATc1_mut_vs_WT_results_dir, "/signifLog2Promoters_DOWN_GOBP_barplot_ntop",ntop_GOTERMS,".png"))

# only top25
ntop_GOTERMS=25

# https://ggplot2.tidyverse.org/reference/geom_bar.html
signifLog2Promoters_DOWN_GOBP_barplot <- ggplot(signifLog2Promoters_DOWN_goEnrichment_GOBP_toPlot[1:ntop_GOTERMS,], aes(x=log10padj, y=fct_rev(Description), fill = log10padj)) + 
  scale_y_discrete(labels = scales::wrap_format(60)) +
  geom_col() +
  theme_bw() +
  xlab("-log10(padj)") + ylab("") +
  labs(fill = "-log10(padj)") + 
  scale_fill_gradient2(low="navy", mid="white", high="red", 
                       midpoint=0)

# width = 10, height = 20, units = "cm"
ggsave(signifLog2Promoters_DOWN_GOBP_barplot, filename = paste0(NFATc1_mut_vs_WT_results_dir, "/signifLog2Promoters_DOWN_GOBP_barplot_ntop",ntop_GOTERMS,".pdf"))
ggsave(signifLog2Promoters_DOWN_GOBP_barplot, filename = paste0(NFATc1_mut_vs_WT_results_dir, "/signifLog2Promoters_DOWN_GOBP_barplot_ntop",ntop_GOTERMS,".png"))

```

```{r plotting IL2, include=TRUE, eval=TRUE, message=FALSE}
# IL2, STAT1
# IL2 - entrez_id: 3558
# - [ ] add peak info - where is the peak!!!
library(karyoploteR)
# is start coming from BiocGenerics?
gene_cols <- c("gene_id", "tx_id", "tx_name", "tx_chrom", "tx_strand")
txdb_genes <- GenomicFeatures::genes(txdb, columns=gene_cols, single.strand.genes.only=TRUE)
IL2_region <- txdb_genes[txdb_genes$gene_id == "3558"]
extend_region <- 5000
start(IL2_region) <- start(IL2_region)  - extend_region  # extending region aroung IL2
end(IL2_region) <- end(IL2_region)  + extend_region
start(IL2_region)
end(IL2_region)

genes.data <- karyoploteR::makeGenesDataFromTxDb(txdb = txdb,
                                    karyoplot=kp,
                                    plot.transcripts = TRUE, 
                                    plot.transcripts.structure = TRUE)

kp <- plotKaryotype(genome = "hg38", zoom = IL2_region, cex=2)
karyoploteR::kpPlotGenes(kp, data=genes.data)

genes.data <- karyoploteR::addGeneNames(genes.data, orgDb="auto")
genes.data <- karyoploteR::mergeTranscripts(genes.data)

pp <- getDefaultPlotParams(plot.type=1)
pp$leftmargin <- 0.15
pp$topmargin <- 15
pp$bottommargin <- 15
pp$ideogramheight <- 5
pp$data1inmargin <- 10
pp$data1outmargin <- 0

gene_toPlot="IL2"

# using .bigWig files generate by BSF
# normalization:  (reads per genome coverage, RPGC).
# 1x (effective genome size 2747877777) 
# bamFilesList: ['/nobackup/lab_bsf/projects/BSA_0612_NFATC1_T_Cells/atacseq_results/PID1185/mapped/PID1185.filtered.bam']
# binLength: 10
# ignoreDuplicates: False
# chrsToSkip: []
samples_bw <- c(PID1185="/home/rstudio/workspace/datasets/atacseq/bam_files/PID1185.bigWig",
                PID1180="/home/rstudio/workspace/datasets/atacseq/bam_files/PID1180.bigWig",
                ND_MK="/home/rstudio/workspace/datasets/atacseq/bam_files/ND_MK.bigWig",
                ND_RJH="/home/rstudio/workspace/datasets/atacseq/bam_files/ND_RJH.bigWig")

kp <- plotKaryotype(zoom = IL2_region, cex=2, genome = "hg38", plot.params = pp)
kpAddBaseNumbers(kp, tick.dist = 10000, minor.tick.dist = 2000,
                add.units = TRUE, cex=2, tick.len = 3)
kpAddMainTitle(kp, paste0("ATAC-seq signal in ", gene_toPlot), cex=2)
kpPlotGenes(kp, data=genes.data, r0=0, r1=0.1, gene.name.cex = 2.5)

#atac-seq signal
total.tracks <- length(samples_bw) #+length(DNA.binding)
out.at <- autotrack(1:length(samples_bw), total.tracks, margin = 0.3, r0=0.23)

# normalized?
# https://rdrr.io/bioc/karyoploteR/man/kpPlotBAMDensity.html
patient_samples=c("PID1185", "PID1180")
ymax_limit=16
for(i in seq_len(length(samples_bw))) {
  bigwig.file <- samples_bw[i]
  at <- autotrack(i, length(samples_bw), r0=out.at$r0, r1=out.at$r1, margin = 0.1)
  if (names(bigwig.file) %in% patient_samples) {
      kp <- kpPlotBigWig(kp, data=bigwig.file, 
                         ymax=ymax_limit,
                         #ymax="visible.region",
                     r0=at$r0, r1=at$r1, col = "red")
  } else {
    kp <- kpPlotBigWig(kp, data=bigwig.file, 
                       ymax=ymax_limit,
                       #ymax="visible.region",
                     r0=at$r0, r1=at$r1, col = "cadetblue2")
  }
    
  computed.ymax <- ceiling(kp$latest.plot$computed.values$ymax)  # 26 - max from C6; fix the scale?!
  kpAxis(kp, ymin=0, 
         ymax=computed.ymax, 
         #tick.pos = c(130),
         tick.pos = computed.ymax, 
         r0=at$r0, r1=at$r1, cex=1.6)
  kpAddLabels(kp, labels = "RPGC", r0=at$r0, r1=at$r1,
              cex=1.0, srt=90, pos=1, label.margin = 0.02)  # 0.14
  kpAddLabels(kp, labels = names(samples_bw)[i], r0=at$r0, r1=at$r1, 
              cex=1.6, label.margin = 0.05)  # 0.035
}


# saving as pdf
pdf(file = file.path(NFATc1_mut_vs_WT_results_dir, paste0("gene_tracks_",gene_toPlot, ".pdf")), width=18, height=10)
kp <- plotKaryotype(zoom = IL2_region, cex=2, genome = "hg38", plot.params = pp)
kpAddBaseNumbers(kp, tick.dist = 10000, minor.tick.dist = 2000,
                add.units = TRUE, cex=2, tick.len = 3)
kpAddMainTitle(kp, paste0("ATAC-seq signal in ", gene_toPlot), cex=2)
kpPlotGenes(kp, data=genes.data, r0=0, r1=0.1, gene.name.cex = 2.5)

#atac-seq signal
total.tracks <- length(samples_bw) #+length(DNA.binding)
out.at <- autotrack(1:length(samples_bw), total.tracks, margin = 0.3, r0=0.23)
for(i in seq_len(length(samples_bw))) {
  bigwig.file <- samples_bw[i]
  at <- autotrack(i, length(samples_bw), r0=out.at$r0, r1=out.at$r1, margin = 0.1)
  if (names(bigwig.file) %in% patient_samples) {
      kp <- kpPlotBigWig(kp, data=bigwig.file, 
                         ymax=ymax_limit,
                         #ymax="visible.region",
                     r0=at$r0, r1=at$r1, col = "red")
  } else {
    kp <- kpPlotBigWig(kp, data=bigwig.file, 
                       ymax=ymax_limit,
                       #ymax="visible.region",
                     r0=at$r0, r1=at$r1, col = "cadetblue2")
  }
    
  computed.ymax <- ceiling(kp$latest.plot$computed.values$ymax)  # 26 - max from C6; fix the scale?!
  kpAxis(kp, ymin=0, 
         ymax=computed.ymax, 
         #tick.pos = c(130),
         tick.pos = computed.ymax, 
         r0=at$r0, r1=at$r1, cex=1.6)
  kpAddLabels(kp, labels = "RPGC", r0=at$r0, r1=at$r1,
              cex=1.0, srt=90, pos=1, label.margin = 0.02)  # 0.14
  kpAddLabels(kp, labels = names(samples_bw)[i], r0=at$r0, r1=at$r1, 
              cex=1.6, label.margin = 0.05)  # 0.035
}

dev.off()

png(filename = file.path(NFATc1_mut_vs_WT_results_dir, paste0("gene_tracks_",gene_toPlot, ".png")))
kp <- plotKaryotype(zoom = IL2_region, cex=2, genome = "hg38", plot.params = pp)
kpAddBaseNumbers(kp, tick.dist = 10000, minor.tick.dist = 2000,
                add.units = TRUE, cex=2, tick.len = 3)
kpAddMainTitle(kp, paste0("ATAC-seq signal in ", gene_toPlot), cex=2)
kpPlotGenes(kp, data=genes.data, r0=0, r1=0.1, gene.name.cex = 2.5)

#atac-seq signal
total.tracks <- length(samples_bw) #+length(DNA.binding)
out.at <- autotrack(1:length(samples_bw), total.tracks, margin = 0.3, r0=0.23)

# normalized?
# https://rdrr.io/bioc/karyoploteR/man/kpPlotBAMDensity.html

for(i in seq_len(length(samples_bw))) {
  bigwig.file <- samples_bw[i]
  at <- autotrack(i, length(samples_bw), r0=out.at$r0, r1=out.at$r1, margin = 0.1)
  if (names(bigwig.file) %in% patient_samples) {
      kp <- kpPlotBigWig(kp, data=bigwig.file, 
                         ymax=ymax_limit,
                         #ymax="visible.region",
                     r0=at$r0, r1=at$r1, col = "red")
  } else {
    kp <- kpPlotBigWig(kp, data=bigwig.file, 
                       ymax=ymax_limit,
                       #ymax="visible.region",
                     r0=at$r0, r1=at$r1, col = "cadetblue2")
  }
    
  computed.ymax <- ceiling(kp$latest.plot$computed.values$ymax)  # 26 - max from C6; fix the scale?!
  kpAxis(kp, ymin=0, 
         ymax=computed.ymax, 
         #tick.pos = c(130),
         tick.pos = computed.ymax, 
         r0=at$r0, r1=at$r1, cex=1.6)
  kpAddLabels(kp, labels = "RPGC", r0=at$r0, r1=at$r1,
              cex=1.0, srt=90, pos=1, label.margin = 0.02)  # 0.14
  kpAddLabels(kp, labels = names(samples_bw)[i], r0=at$r0, r1=at$r1, 
              cex=1.6, label.margin = 0.05)  # 0.035
}

dev.off()

```


```{r finding Motifs, include=FALSE, eval=FALSE, cache=TRUE}
#genes_interest <- c("IKZF1", "IKZF2", "NFAT5", "NFATC1", "NFATC2", "NFATC3", "NFATC4", "NFKB")

# IKZF1 and IKZF2
# IKZF1 <- query(MotifDb, c("IKZF1")) 
# IKZF1_IKAROS <- query(MotifDb, c("IKAROS")) # ?? one match points to one of the above resource: Hsapiens-HOCOMOCOv11-core-C-IKZF1_HUMAN.H11MO.0.C
# plot motifs for different NFATC(N)
#NFATC1 <- TFBSTools::query(motifsToScan, c("NFATC1")) 
# # IKZF2_HELIOS <- query(MotifDb, c("HELIOS")) # NULL
# JUN <- query(MotifDb, c("JUN"))  #  there are multiple JUN entries
# 
# names(IKZF2)
# 
# ikzf1_jaspar_Motif <- IKZF1[["Hsapiens-HOCOMOCOv10-IKZF1_HUMAN.H10MO.C"]]
# ikzf1_Motif <- IKZF1[["Hsapiens-SwissRegulon-IKZF1.SwissRegulon"]]
# ikzf2_Motif <- IKZF2[["Hsapiens-SwissRegulon-IKZF2.SwissRegulon"]]
# 
# ikzf1_Motif[, 1:4]
# ikzf2_Motif[, 1:4]
# 
# seqLogo(ikzf1_jaspar_Motif)
# seqLogo(ikzf1_Motif)
# seqLogo(ikzf2_Motif)

# TFBSTools for motif sets

# Identifying Motifs in ATACs-seq ---
#First we can retrieve a sensible set of motifs to scan for within our mouse tissue ATACseq data.
#Here we retrieve the vertebrate, JASPAR CORE motifs. We additional specify all_versions is FALSE to only include the latest version of a motif.

if (!file.exists(paste0(results_dir, "/motif_enrichment.RData"))) {
  #TODO:
  # [ ] - check effect of normalization?
  
  #?getMatrixSet
  opts <- list()
  opts[["tax_group"]] <- "vertebrates"
  opts[["species"]] <- "9606"
  opts[["collection"]] <- "CORE"
  opts[["all_versions"]] <- FALSE
  motifsToScan <- TFBSTools::getMatrixSet(JASPAR2020, opts)
  
  # summarizing ATAC signal to Motifs ----
  # The chromVar package allows for the summarization of ATACseq signal changes to the motifs within peaks.
  # With this summarization we can potentially identify which motifs may have an important role in a set of ATAC samples compared to other samples.
  # The chromVar package comes from the same lab and author as the motifmatchr package and so will work well together.
  
  # Setting up for chromVar ----
  #To identify motifs for chomVar we will use motifMatchr with a different set of inputs.
  #Here we will provide our RangedSummarizedExperiment object contaitng our counts over peaks directly to the matchMotifs function.

  #First we will remove any peaks with less than 5 reads across all samples.
  SummarizedExperiment::colData(counts_consensus) <- SummarizedExperiment::colData(atacDDS)
  
  # using normalized counts from DESeq2
  #counts(counts_consensus) <- atacDDS_normCounts
  
  dim(counts_consensus)
  counts_consensus_filt <- counts_consensus[rowSums(assay(counts_consensus)) > 5, ]
  dim(counts_consensus)
  
  #Correct GC bias ---
  
  #Next we can correct for any potential GC bias which may have arisen in sequencing. 
  # - there seems to be no GC bias so probably not needed. But check initial QC using also ChIPQC packageS?
  
  #For comparing across differing sets of peaks with differing sequences compositions we will need to correct for this bias. We can use the addGCBias and specify our genome as a BSgenome object to correct in chromVar.
  # Computes GC content for peaks
  # rather use fasta file - https://greenleaflab.github.io/chromVAR/articles/Introduction.html#visualizing-deviations
  # or DNAStringSet 

  # added extra
  # counts_consensus_filt <- chromVAR::filterPeaks(counts_consensus_filt, non_overlapping = TRUE)
  # dim(counts_consensus_filt)  # makes no difference
  
  # added extra
  #fastq_file <- "/home/peter_r/resources/hg38.fa"
  #fasta_ref <- Rsamtools::FaFile(fastq_file, index=sprintf("%s.fai", fastq_file))
  #counts_consensus_filt <- chromVAR::addGCBias(counts_consensus_filt, genome = fasta_ref)  # or rather use FaFile!
  #B <- chromVAR::addGCBias(counts_consensus_filt, genome = fasta_ref)
  
  #library(BSgenome.Hsapiens.UCSC.hg38)
  counts_consensus_filt <- chromVAR::addGCBias(counts_consensus_filt, genome = BSgenome.Hsapiens.UCSC.hg38)  # or rather use FaFile!
  #SummarizedExperiment::rowData(counts_consensus_filt)$bias
  #A <- chromVAR::addGCBias(counts_consensus_filt, genome = BSgenome.Hsapiens.UCSC.hg38)
  #A_bias <- rowData(A)$bias #head(rowData(A))
  #B_bias <- rowData(B)$bias #head(rowData(A))
  #identical(A_bias, B_bias) TRUE
  # ID motifs in peaks ----
  
  # Having corrected for bias, we can use the matchMotifs function again to identify motifs under our ATACseq peaks.
  # Here we supply our RangedSummarizedExperiment of counts in peaks and the genome of interest to the matchMotifs function and use the default out of matches.
  
  # added extra - not used!
  # but this fetches JASPAR2016!?
  #motifsToScan <- chromVAR::getJasparMotifs(species = "Homo sapiens", collection = "CORE") # instead of using the above function; compare!
  
  # BSgenome.Hsapiens.UCSC.hg38 compare to NCBI sequence used in the PEPATAC
  # identical with NCBI fasta used in PEPATAC when genome = fasta_ref
  motif_matches <- motifmatchr::matchMotifs(pwms = motifsToScan, 
                                            subject = counts_consensus_filt, 
                                            genome = BSgenome.Hsapiens.UCSC.hg38, 
                                            out = "matches")
  motif_matches
  
  # Run chromVar ----
  
  #Following the identification of motifs in our peaks, we can perform the summarization of ATACseq signal to motifs using the computeDeviations and the computeVariability functions.
  
  # Computes deviations in chromatin accessibility across sets of annotations
  # https://bioconductor.org/packages/release/bioc/vignettes/chromVAR/inst/doc/Introduction.html
  # library(BiocParallel)
  BiocParallel::register(BiocParallel::MulticoreParam(8, progressbar = FALSE))
  set.seed(2021) # !!! seed needs to be set otherwise results differ !!! 
  #register(SerialParam())
  
  # deviations
  # The function computeDeviations returns a SummarizedExperiment with two assays. The first matrix (accessible via deviations(dev) or assays(dev)$deviations) will give the bias corrected deviation in accessibility for each set of peaks (rows) for each cell or sample (columns). This metric represent how accessible the set of peaks is relative to the expectation based on equal chromatin accessibility profiles across cells/samples, normalized by a set of background peak sets matched for GC and average accessability. The second matrix (deviationScores(dev) or assays(deviations)$z) gives the deviation Z-score, which takes into account how likely such a score would occur if randomly sampling sets of beaks with similar GC content and average accessibility.
  
  # The deviations result contains a SummarizedExperiment object with the Z-scores showing the enrichment for ATACseq signal in each sample for every motif.
  
  # Computes deviations in chromatin accessibility across sets of annotations
  # - [ ] what it actually calculates??? see details!
  
  # added extra
  # per-compute background peaks only on middle bit of overlapping peaks
  # or potentially use normalized counts from DESeq?! - why are normalized counts not used by default???!
  # adde extra 
  #  https://greenleaflab.github.io/chromVAR/articles/Articles/Deviations.html
  # use group info to calculate expectations!
  background_peaks <- chromVAR::getBackgroundPeaks(object = counts_consensus_filt) 
  # expectation below seems to produce same results? all.equal() TRUE
  access_expectation <- chromVAR::computeExpectations(object = counts_consensus_filt)  # default
  #access_expectation <- chromVAR::computeExpectations(object = counts_consensus_filt, norm = TRUE) 
  # Warning message:
  # In if (norm) { :
  #  the condition has length > 1 and only the first element will be used
   # access_expectation <- chromVAR::computeExpectations(object = counts_consensus_filt, group = colData(counts_consensus_filt)$sample_group) # at least two samples for normalization!
  #access_expectation <- chromVAR::computeExpectations(object = counts_consensus_filt, colData(counts_consensus_filt)$sample_group) 
  #access_expectation <- chromVAR::computeExpectations(object = counts_consensus_filt, colData(counts_consensus_filt)$sample_group, norm = TRUE) # this or group = "sample_group" are identical!

  chrom_access_deviations <- chromVAR::computeDeviations(object = counts_consensus_filt, 
                                                         annotations = motif_matches,
                                                         background_peaks = background_peaks,
                                                         expectation = access_expectation)
  # adding samples names
  #colData(chrom_access_deviations)
  #colData(counts_consensus_filt)
  
  #colData(chrom_access_deviations) <- colData(atacDDS)  # check if order of samples is the same between chrom_access_deviations and atacDDS
  
  # plotVariability(variability, use_plotly = FALSE) 
  sample_cor <- chromVAR::getSampleCorrelation(chrom_access_deviations)
  
  #library(pheatmap)
  annotation_row_sampleCor <- as.data.frame(SummarizedExperiment::colData(chrom_access_deviations)) %>%
    dplyr::select(genotype)
  pheatmap::pheatmap(as.dist(sample_cor), 
                     annotation_row = annotation_row_sampleCor, 
                     clustering_distance_rows = as.dist(1-sample_cor), 
                     clustering_distance_cols = as.dist(1-sample_cor), 
                     filename = paste0(NFATc1_mut_vs_WT_results_motif_dir, "/chrom_access_deviations_sample_corr.pdf"))
  
  # assessing similarity
  tsne_results <- chromVAR::deviationsTsne(chrom_access_deviations, threshold = 1.5, perplexity = 8, 
                                what = "samples", shiny = FALSE)
  #ggplot(tsne_results, aes(x = Dim1, y = Dim2)) + geom_point() + 
  #  chromVAR::chromVAR_theme()
  # annotation = "IL2", 
  # check genes for annotation - View(as.data.frame(SummarizedExperiment::rowData(chrom_access_deviations)))
  tsne_plots <- chromVAR::plotDeviationsTsne(chrom_access_deviations, tsne_results,
                                             annotation = c("NFATC2", "NFATC3", "NFATC4"),
                                    sample_column = "genotype", shiny = FALSE)
  # tsne_plots[[1]]
  # tsne_plots[[2]]
  #renv::install("ggpubr")
  ggpubr::ggexport(plotlist = tsne_plots, 
                   filename = paste0(NFATc1_mut_vs_WT_results_motif_dir, "/chrom_access_deviations_sample_tsne.pdf"))
  

  # Motif / kmer similarity
  inv_tsne_results <- chromVAR::deviationsTsne(chrom_access_deviations, threshold = 1.5, perplexity = 8, 
                                    what = "annotations", shiny = FALSE)
  
  # test from 
  # https://github.com/GreenleafLab/chromVAR/issues/32
  
  data.frame(chrom_access_variability) %>% dplyr::arrange(desc(variability))-> motifVariability
  motifVariability$rank <- 1:dim(chrom_access_variability)[1]
  ggplot(motifVariability, aes(x = rank, y = variability)) + geom_point(size = 0.75) +
    labs(x = "Motif Rank", y = "Accessibility Variability")
  
  # check the following for peak selection:
  # https://github.com/GreenleafLab/chromVAR/issues/81
  # https://github.com/GreenleafLab/ArchR
  # how to proceed next - e.g. motif identification
  motif_similarity_tsne <- ggplot(inv_tsne_results, aes(x = Dim1, y = Dim2)) + geom_point() + 
    chromVAR::chromVAR_theme()
    
  # Differential accessibility and variability of motifs!
  # https://greenleaflab.github.io/chromVAR/articles/Articles/Applications.html
  diff_acc <- chromVAR::differentialDeviations(chrom_access_deviations, groups="genotype", parametric = FALSE)
  motif_annot <- as.data.frame(SummarizedExperiment::rowData(chrom_access_deviations)) %>%
    tibble::rownames_to_column(var = "motif") %>%
    dplyr::select(motif, name)
  diff_acc_annot <- diff_acc %>%
    tibble::rownames_to_column(var = "motif") %>%
    dplyr::left_join(., motif_annot, by = "motif")
  head(diff_acc_annot)  ## parametric = TRUE not enough 'y' observations
  
  diff_var <- chromVAR::differentialVariability(chrom_access_deviations, "genotype", parametric = TRUE)
  diff_var_annot <- diff_var %>%
    tibble::rownames_to_column(var = "motif") %>%
    dplyr::left_join(., motif_annot, by = "motif")
  head(diff_var_annot)

  devZscores <- chromVAR::deviationScores(chrom_access_deviations)
  devZscores_df <- tibble::rownames_to_column(as.data.frame(devZscores), var="jaspar_id")
  
  # function to compute overall variability of motif sets across samples ----
  chrom_access_variability <- chromVAR::computeVariability(chrom_access_deviations)
  chrom_access_variability_plot <- chromVAR::plotVariability(chrom_access_variability, use_plotly = FALSE)
  ggsave(chrom_access_variability_plot, filename = paste0(NFATc1_mut_vs_WT_results_motif_dir, "/chrom_access_variability.pdf"))
  
  BiocParallel::register(BiocParallel::SerialParam())

  chrom_access_variability_ord <- chrom_access_variability[order(chrom_access_variability$p_value), ]  # ordering based on p-value
  #chrom_access_variability_ord[1:10, ]

  #! to store
  chrom_access_variability_ord_results <- chrom_access_variability_ord %>%
    tibble::rownames_to_column(var="jaspar_id") %>%
    dplyr::left_join(., devZscores_df, by = "jaspar_id")
  
  # both patient and normals going change in the same direction
  chrom_access_variability_ord_results_directions <- chrom_access_variability_ord_results %>%
    dplyr::filter(p_value_adj < 0.05) %>%
    dplyr::mutate(nd_same_direction = dplyr::if_else(sign(ND_RJH) == sign(ND_MK), TRUE, FALSE),
                  pat_same_direction = dplyr::if_else(sign(PID1185) == sign(PID1180), TRUE, FALSE)) 
  
  chrom_access_variability_ord_results_directions_signifConsistPatients <- chrom_access_variability_ord_results_directions %>%
    dplyr::filter(nd_same_direction == TRUE & pat_same_direction == TRUE)
  
  #sum(chrom_access_variability_ord_results$p_value_adj < 0.05)
  
  #plotVariability(chrom_access_variability, use_plotly = FALSE) 
  
  message("Saving motif enrichment results")
  save(counts_consensus, chrom_access_deviations, chrom_access_variability, chrom_access_variability_ord, chrom_access_variability_ord_results, file = paste0(NFATc1_mut_vs_WT_results_motif_dir, "/motif_enrichment.RData"))
} else {
  
  message("Loading motif enrichment results")
  load(file = paste0(NFATc1_mut_vs_WT_results_motif_dir, "/motif_enrichment.RData"))
  
  devZscores <- chromVAR::deviationScores(chrom_access_deviations)
  devZscores_df <- tibble::rownames_to_column(as.data.frame(devZscores), var="jaspar_id")
  
}

# chromVar results ----

# We can use the results from the variability with our Z-score deviations to identify in which sample our motifs are enriched.
# chromVAR::computeVariability
ntop <- 50
#FIXME:
# change chrom_access_variability_ord ->  chrom_access_variability_ord_signif in topVariable <- chrom_access_variability_ord[1:ntop, ]
chrom_access_variability_ord_signif <- chrom_access_variability_ord %>%
  dplyr::filter(p_value_adj < motif_padj_cutoff)
topVariable <- chrom_access_variability_ord_signif[1:ntop, ]
topVariable <- tibble::rownames_to_column(topVariable, var="jaspar_id")

#devTop <- merge(topVariable[, 1, drop = FALSE], devZscores, by = 0)  # by = 0 - rownames; but gives random order?!
devTop <- topVariable %>%
  dplyr::left_join(., devZscores_df, by = "jaspar_id")
#devTop[1:2, ]

# A useful way to visualize these results is using a heatmap. Although we cover this in a later session, here we can use the pheatmap library with default settings to illustrate where our most variable motifs are active.
#devToPlot <- as.matrix(devTop[, -c(1:2)])
devToPlot <- devTop %>%
  dplyr::select(name, ND_RJH, ND_MK, PID1180, PID1185) %>%
  tibble::column_to_rownames(var = "name")
#rownames(devToPlot) <- devTop[, 2]
#library(pheatmap)
annotCol_forHeatmap <- project_metadata %>%
  dplyr::select(sample_name, genotype) %>%
  tibble::column_to_rownames(var="sample_name")

pheatmap::pheatmap(devToPlot, 
                   scale="none", 
                   cluster_cols = FALSE, 
                   annotation_col = annotCol_forHeatmap, 
                   main = paste0("Heatmap of motif enrichment - top ", ntop))
pheatmap::pheatmap(devToPlot,
                   scale = "none",
                   cluster_cols = FALSE,
                   annotation_col = annotCol_forHeatmap,
                   main = paste0("Heatmap of motif enrichment - top ", ntop),
                   filename = file.path(NFATc1_mut_vs_WT_results_motif_dir, paste0("motif_enrichment_signifTop", ntop,".pdf")))
# 
# pheatmap::pheatmap(devToPlot, 
#                    scale = "none", 
#                    cluster_cols = FALSE,
#                    annotation_col = annotCol_forHeatmap,
#                    main = paste0("Heatmap of motif enrichment - top ", ntop),
#                    filename = file.path(out_dir, paste0("motif_enrichment_signifTop", ntop,".png")))
#motif_dev_heatmap <- ComplexHeatmap::pheatmap(as.matrix(devToPlot), scale="none", cluster_cols = FALSE)

# replacing heatmap with publication heatmap ----
library("RColorBrewer")
annotCol_forHeatmap_colors <- list(genotype = c(  WT="#525252", `NFATc1_mut` = "#fc4e2a"),
                                   bio_rep = c(`ND_RJH` = "#d9d9d9", `ND_MK` = "#969696", `PID1185` = "#fd8d3c", `PID1180` = "#fd8d3c"))
#motifs_heatmap <-  
pheatmap::pheatmap(as.matrix(devToPlot),
                   color = colorRampPalette(rev(brewer.pal(7, "PiYG")))(100),
                   scale = "row",
                   #gaps_col = ph_col_annot%>% arrange(state)%$% table(state) %>% cumsum,
                   #gaps_row = ph_genes$clust %>% table() %>% cumsum,
                   cluster_cols = FALSE,
                   cluster_rows = TRUE,
                   show_colnames = FALSE,
                   show_rownames = TRUE,
                   annotation_col = annotCol_forHeatmap,
                   #annotation_row = ph_row_annot,
                   #annotation_legend = FALSE,
                   annotation_colors = annotCol_forHeatmap_colors,  
                   #labels_row = row_labels,
                   fontsize_row = 9,
                   fontsize = 9)

pheatmap::pheatmap(as.matrix(devToPlot),
                   color = colorRampPalette(rev(brewer.pal(7, "PiYG")))(100),
                   scale = "row",
                   #gaps_col = ph_col_annot%>% arrange(state)%$% table(state) %>% cumsum,
                   #gaps_row = ph_genes$clust %>% table() %>% cumsum,
                   cluster_cols = FALSE,
                   cluster_rows = TRUE,
                   show_colnames = FALSE,
                   show_rownames = TRUE,
                   annotation_col = annotCol_forHeatmap,
                   #annotation_row = ph_row_annot,
                   #annotation_legend = FALSE,
                   annotation_colors = annotCol_forHeatmap_colors,  
                   #labels_row = row_labels,
                   fontsize_row = 9,
                   fontsize = 9,
                   #cellheight = 9,
                   main = paste0("Heatmap of motif enrichment - top ", ntop),
                   filename = file.path(NFATc1_mut_vs_WT_results_motif_dir, paste0("motif_enrichment_signifTop", ntop,".pdf")))

pheatmap::pheatmap(as.matrix(devToPlot),
                   color = colorRampPalette(rev(brewer.pal(7, "PiYG")))(100),
                   scale = "row",
                   #gaps_col = ph_col_annot%>% arrange(state)%$% table(state) %>% cumsum,
                   #gaps_row = ph_genes$clust %>% table() %>% cumsum,
                   cluster_cols = FALSE,
                   cluster_rows = TRUE,
                   show_colnames = FALSE,
                   show_rownames = TRUE,
                   annotation_col = annotCol_forHeatmap,
                   #annotation_row = ph_row_annot,
                   #annotation_legend = FALSE,
                   annotation_colors = annotCol_forHeatmap_colors,  
                   #labels_row = row_labels,
                   fontsize_row = 9,
                   fontsize = 9,
                   #cellheight = 9,
                   main = paste0("Heatmap of motif enrichment - top ", ntop),
                   filename = file.path(NFATc1_mut_vs_WT_results_motif_dir, paste0("motif_enrichment_signifTop", ntop,".png")))

# pdf(file.path(out_dir, paste0("motif_enrichment_signifTop", ntop,".pdf")))
# grid.draw(motifs_heatmap)
# dev.off()

#TODO:
# do actuall stats test that there is significant different in variability between my group of interest
motif_results_list <- list(motif_variab_raw = chrom_access_variability_ord_results,
                           motif_variab_signif = chrom_access_variability_ord_results_directions,
                           motif_variab_signif_direction = chrom_access_variability_ord_results_directions_signifConsistPatients)
openxlsx::write.xlsx(x = motif_results_list, 
                     file = paste0(NFATc1_mut_vs_WT_results_motif_dir, "/motif_enrichment.xlsx"),
                     asTable = TRUE,
                     overwrite = TRUE) 


```

```{r finding Motifs - all, include=FALSE, eval=FALSE, cache=TRUE}
# Running for all peaks; not just consensus!!!

if (!file.exists(paste0(results_dir, "/motif_enrichment.RData"))) {
  #?getMatrixSet
  opts <- list()
  opts[["tax_group"]] <- "vertebrates"
  opts[["species"]] <- "9606"
  opts[["collection"]] <- "CORE"
  opts[["all_versions"]] <- FALSE
  motifsToScan <- TFBSTools::getMatrixSet(JASPAR2020, opts)
  
  # summarizing ATAC signal to Motifs ----
  # The chromVar package allows for the summarization of ATACseq signal changes to the motifs within peaks.
  # With this summarization we can potentially identify which motifs may have an important role in a set of ATAC samples compared to other samples.
  # The chromVar package comes from the same lab and author as the motifmatchr package and so will work well together.
  
  # Setting up for chromVar ----
  #To identify motifs for chomVar we will use motifMatchr with a different set of inputs.
  #Here we will provide our RangedSummarizedExperiment object contaitng our counts over peaks directly to the matchMotifs function.
    bamFL <- Rsamtools::BamFileList(project_metadata$bam_files_path, yieldSize = 5e+06)
    counts_all_peaks <- GenomicAlignments::summarizeOverlaps(features = nrToCount, 
                                                           reads = bamFL, 
                                                           singleEnd = TRUE,
                                                           ignore.strand = TRUE,
                                                           fragments = FALSE,
                                                           param = Rsamtools::ScanBamParam(flag = Rsamtools::scanBamFlag(isDuplicate = FALSE)))
  
  dim(counts_all_peaks)
  counts_filtered <- chromVAR::filterPeaks(counts_all_peaks, non_overlapping = TRUE, min_fragments_per_peak = 10)
  dim(counts_filtered)
  
  # testing using all peaks
  # AAA <- counts_consensus
  # counts_consensus <- counts_filtered
  
  #First we will remove any peaks with less than 5 reads across all samples.
  SummarizedExperiment::colData(counts_filtered) <- SummarizedExperiment::colData(atacDDS)
  
  # using normalized counts from DESeq2
  # not using normalized counts!!! - there is no normalization for since peaks are not shared!?
  #TODO:
  # [ ] - check with/without normalization for consensus peaks (run above!)
  #counts(counts_consensus) <- atacDDS_normCounts
  
  #Correct GC bias ---
  
  #Next we can correct for any potential GC bias which may have arisen in sequencing. 
  # - there seems to be no GC bias so probably not needed. But check initial QC using also ChIPQC packageS?
  
  #For comparing across differing sets of peaks with differing sequences compositions we will need to correct for this bias. We can use the addGCBias and specify our genome as a BSgenome object to correct in chromVar.
  # Computes GC content for peaks
  # rather use fasta file - https://greenleaflab.github.io/chromVAR/articles/Introduction.html#visualizing-deviations
  # or DNAStringSet 

  # added extra
  # counts_consensus_filt <- chromVAR::filterPeaks(counts_consensus_filt, non_overlapping = TRUE)
  # dim(counts_consensus_filt)  # makes no difference
  
  # added extra
  #fastq_file <- "/home/peter_r/resources/hg38.fa"
  #fasta_ref <- Rsamtools::FaFile(fastq_file, index=sprintf("%s.fai", fastq_file))
  #counts_consensus_filt <- chromVAR::addGCBias(counts_consensus_filt, genome = fasta_ref)  # or rather use FaFile!
  #B <- chromVAR::addGCBias(counts_consensus_filt, genome = fasta_ref)
  
  #library(BSgenome.Hsapiens.UCSC.hg38)
  counts_filtered <- chromVAR::addGCBias(counts_filtered, genome = BSgenome.Hsapiens.UCSC.hg38)  # or rather use FaFile!
  #SummarizedExperiment::rowData(counts_consensus_filt)$bias
  #A <- chromVAR::addGCBias(counts_consensus_filt, genome = BSgenome.Hsapiens.UCSC.hg38)
  #A_bias <- rowData(A)$bias #head(rowData(A))
  #B_bias <- rowData(B)$bias #head(rowData(A))
  #identical(A_bias, B_bias) TRUE
  # ID motifs in peaks ----
  
  # Having corrected for bias, we can use the matchMotifs function again to identify motifs under our ATACseq peaks.
  # Here we supply our RangedSummarizedExperiment of counts in peaks and the genome of interest to the matchMotifs function and use the default out of matches.
  
  # added extra - not used!
  # but this fetches JASPAR2016!?
  #motifsToScan <- chromVAR::getJasparMotifs(species = "Homo sapiens", collection = "CORE") # instead of using the above function; compare!
  
  # BSgenome.Hsapiens.UCSC.hg38 compare to NCBI sequence used in the PEPATAC
  # identical with NCBI fasta used in PEPATAC when genome = fasta_ref
  motif_matches <- motifmatchr::matchMotifs(pwms = motifsToScan, 
                                            subject = counts_filtered, 
                                            genome = BSgenome.Hsapiens.UCSC.hg38, 
                                            out = "matches")
  motif_matches
  
  # Run chromVar ----
  
  #Following the identification of motifs in our peaks, we can perform the summarization of ATACseq signal to motifs using the computeDeviations and the computeVariability functions.
  
  # Computes deviations in chromatin accessibility across sets of annotations
  # https://bioconductor.org/packages/release/bioc/vignettes/chromVAR/inst/doc/Introduction.html
  # library(BiocParallel)
  BiocParallel::register(BiocParallel::MulticoreParam(8, progressbar = FALSE))
  set.seed(2021) # !!! seed needs to be set otherwise results differ !!! 
  #register(SerialParam())
  
  # deviations
  # The function computeDeviations returns a SummarizedExperiment with two assays. The first matrix (accessible via deviations(dev) or assays(dev)$deviations) will give the bias corrected deviation in accessibility for each set of peaks (rows) for each cell or sample (columns). This metric represent how accessible the set of peaks is relative to the expectation based on equal chromatin accessibility profiles across cells/samples, normalized by a set of background peak sets matched for GC and average accessability. The second matrix (deviationScores(dev) or assays(deviations)$z) gives the deviation Z-score, which takes into account how likely such a score would occur if randomly sampling sets of beaks with similar GC content and average accessibility.
  
  # The deviations result contains a SummarizedExperiment object with the Z-scores showing the enrichment for ATACseq signal in each sample for every motif.
  
  # Computes deviations in chromatin accessibility across sets of annotations
  # - [ ] what it actually calculates??? see details!
  
  # added extra
  # per-compute background peaks only on middle bit of overlapping peaks
  # or potentially use normalized counts from DESeq?! - why are normalized counts not used by default???!
  # adde extra 
  #  https://greenleaflab.github.io/chromVAR/articles/Articles/Deviations.html
  # use group info to calculate expectations!
  background_peaks <- chromVAR::getBackgroundPeaks(object = counts_filtered) 
  # expectation below seems to produce same results? all.equal() TRUE
  access_expectation <- chromVAR::computeExpectations(object = counts_filtered)  # default
  #access_expectation <- chromVAR::computeExpectations(object = counts_consensus_filt, norm = TRUE) 
  # Warning message:
  # In if (norm) { :
  #  the condition has length > 1 and only the first element will be used
   # access_expectation <- chromVAR::computeExpectations(object = counts_consensus_filt, group = colData(counts_consensus_filt)$sample_group) # at least two samples for normalization!
  #access_expectation <- chromVAR::computeExpectations(object = counts_consensus_filt, colData(counts_consensus_filt)$sample_group) 
  #access_expectation <- chromVAR::computeExpectations(object = counts_consensus_filt, colData(counts_consensus_filt)$sample_group, norm = TRUE) # this or group = "sample_group" are identical!

  chrom_access_deviations <- chromVAR::computeDeviations(object = counts_filtered, 
                                                         annotations = motif_matches,
                                                         background_peaks = background_peaks,
                                                         expectation = access_expectation)
  # adding samples names
  #colData(chrom_access_deviations)
  #colData(counts_consensus_filt)
  
  #colData(chrom_access_deviations) <- colData(atacDDS)  # check if order of samples is the same between chrom_access_deviations and atacDDS
  
  # plotVariability(variability, use_plotly = FALSE) 
  sample_cor <- chromVAR::getSampleCorrelation(chrom_access_deviations)
  
  #library(pheatmap)
  annotation_row_sampleCor <- as.data.frame(SummarizedExperiment::colData(chrom_access_deviations)) %>%
    dplyr::select(genotype)
  pheatmap::pheatmap(as.dist(sample_cor), 
                     annotation_row = annotation_row_sampleCor, 
                     clustering_distance_rows = as.dist(1-sample_cor), 
                     clustering_distance_cols = as.dist(1-sample_cor), 
                     filename = paste0(NFATc1_mut_vs_WT_results_motif_allPeaks_dir, "/chrom_access_deviations_sample_corr.pdf"))
  
  # assessing similarity
  tsne_results <- chromVAR::deviationsTsne(chrom_access_deviations, threshold = 1.5, perplexity = 8, 
                                what = "samples", shiny = FALSE)
  #ggplot(tsne_results, aes(x = Dim1, y = Dim2)) + geom_point() + 
  #  chromVAR::chromVAR_theme()
  # annotation = "IL2", 
  # check genes for annotation - View(as.data.frame(SummarizedExperiment::rowData(chrom_access_deviations)))
  tsne_plots <- chromVAR::plotDeviationsTsne(chrom_access_deviations, tsne_results,
                                             annotation = c("NFATC2", "NFATC3", "NFATC4"),
                                    sample_column = "genotype", shiny = FALSE)
  # tsne_plots[[1]]
  # tsne_plots[[2]]
  #renv::install("ggpubr")
  ggpubr::ggexport(plotlist = tsne_plots, 
                   filename = paste0(NFATc1_mut_vs_WT_results_motif_allPeaks_dir, "/chrom_access_deviations_sample_tsne.pdf"))
  

  # Motif / kmer similarity
  inv_tsne_results <- chromVAR::deviationsTsne(chrom_access_deviations, threshold = 1.5, perplexity = 8, 
                                    what = "annotations", shiny = FALSE)
  
  # test from 
  # https://github.com/GreenleafLab/chromVAR/issues/32
  
  data.frame(chrom_access_variability) %>% dplyr::arrange(desc(variability))-> motifVariability
  motifVariability$rank <- 1:dim(chrom_access_variability)[1]
  ggplot(motifVariability, aes(x = rank, y = variability)) + geom_point(size = 0.75) +
    labs(x = "Motif Rank", y = "Accessibility Variability")
  
  # check the following for peak selection:
  # https://github.com/GreenleafLab/chromVAR/issues/81
  # https://github.com/GreenleafLab/ArchR
  # how to proceed next - e.g. motif identification
  motif_similarity_tsne <- ggplot(inv_tsne_results, aes(x = Dim1, y = Dim2)) + geom_point() + 
    chromVAR::chromVAR_theme()
    
  # Differential accessibility and variability of motifs!
  # https://greenleaflab.github.io/chromVAR/articles/Articles/Applications.html
  diff_acc <- chromVAR::differentialDeviations(chrom_access_deviations, groups="genotype", parametric = FALSE)
  motif_annot <- as.data.frame(SummarizedExperiment::rowData(chrom_access_deviations)) %>%
    tibble::rownames_to_column(var = "motif") %>%
    dplyr::select(motif, name)
  diff_acc_annot <- diff_acc %>%
    tibble::rownames_to_column(var = "motif") %>%
    dplyr::left_join(., motif_annot, by = "motif")
  head(diff_acc_annot)  ## parametric = TRUE not enough 'y' observations
  
  diff_var <- chromVAR::differentialVariability(chrom_access_deviations, "genotype", parametric = FALSE)
  diff_var_annot <- diff_var %>%
    tibble::rownames_to_column(var = "motif") %>%
    dplyr::left_join(., motif_annot, by = "motif")
  head(diff_var_annot)

  devZscores <- chromVAR::deviationScores(chrom_access_deviations)
  devZscores_df <- tibble::rownames_to_column(as.data.frame(devZscores), var="jaspar_id")
  
  # function to compute overall variability of motif sets across samples ----
  chrom_access_variability <- chromVAR::computeVariability(chrom_access_deviations)
  chrom_access_variability_plot <- chromVAR::plotVariability(chrom_access_variability, use_plotly = FALSE)
  ggsave(chrom_access_variability_plot, filename = paste0(NFATc1_mut_vs_WT_results_motif_allPeaks_dir, "/chrom_access_variability.pdf"))
  
  BiocParallel::register(BiocParallel::SerialParam())

  chrom_access_variability_ord <- chrom_access_variability[order(chrom_access_variability$p_value), ]  # ordering based on p-value
  #chrom_access_variability_ord[1:10, ]

  #! to store
  chrom_access_variability_ord_results <- chrom_access_variability_ord %>%
    tibble::rownames_to_column(var="jaspar_id") %>%
    dplyr::left_join(., devZscores_df, by = "jaspar_id")
  
  # both patient and normals going change in the same direction
  chrom_access_variability_ord_results_directions <- chrom_access_variability_ord_results %>%
    dplyr::filter(p_value_adj < 0.05) %>%
    dplyr::mutate(nd_same_direction = dplyr::if_else(sign(ND_RJH) == sign(ND_MK), TRUE, FALSE),
                  pat_same_direction = dplyr::if_else(sign(PID1185) == sign(PID1180), TRUE, FALSE)) 
  
  chrom_access_variability_ord_results_directions_signifConsistPatients <- chrom_access_variability_ord_results_directions %>%
    dplyr::filter(nd_same_direction == TRUE & pat_same_direction == TRUE)
  
  #sum(chrom_access_variability_ord_results$p_value_adj < 0.05)
  
  #plotVariability(chrom_access_variability, use_plotly = FALSE) 
  
  message("Saving motif enrichment results")
  save(counts_filtered, chrom_access_deviations, chrom_access_variability, chrom_access_variability_ord, chrom_access_variability_ord_results, file = paste0(NFATc1_mut_vs_WT_results_motif_allPeaks_dir, "/motif_enrichment_allPeaks.RData"))
} else {
  
  message("Loading motif enrichment results")
  load(file = paste0(NFATc1_mut_vs_WT_results_motif_allPeaks_dir, "/motif_enrichment_allPeaks.RData"))
  
  devZscores <- chromVAR::deviationScores(chrom_access_deviations)
  devZscores_df <- tibble::rownames_to_column(as.data.frame(devZscores), var="jaspar_id")
  
}

# chromVar results ----

# We can use the results from the variability with our Z-score deviations to identify in which sample our motifs are enriched.
# chromVAR::computeVariability
ntop <- 50
topVariable <- chrom_access_variability_ord[1:ntop, ]
topVariable <- tibble::rownames_to_column(topVariable, var="jaspar_id")

#devTop <- merge(topVariable[, 1, drop = FALSE], devZscores, by = 0)  # by = 0 - rownames; but gives random order?!
devTop <- topVariable %>%
  dplyr::left_join(., devZscores_df, by = "jaspar_id")
#devTop[1:2, ]

# A useful way to visualize these results is using a heatmap. Although we cover this in a later session, here we can use the pheatmap library with default settings to illustrate where our most variable motifs are active.
#devToPlot <- as.matrix(devTop[, -c(1:2)])
devToPlot <- devTop %>%
  dplyr::select(name, ND_RJH, ND_MK, PID1180, PID1185) %>%
  tibble::column_to_rownames(var = "name")
#rownames(devToPlot) <- devTop[, 2]
#library(pheatmap)
annotCol_forHeatmap <- project_metadata %>%
  dplyr::select(sample_name, genotype) %>%
  tibble::column_to_rownames(var="sample_name")

pheatmap::pheatmap(devToPlot, 
                   scale="none", 
                   cluster_cols = FALSE, 
                   annotation_col = annotCol_forHeatmap, 
                   main = paste0("Heatmap of motif enrichment - top ", ntop))
pheatmap::pheatmap(devToPlot,
                   scale = "none",
                   cluster_cols = FALSE,
                   annotation_col = annotCol_forHeatmap,
                   main = paste0("Heatmap of motif enrichment - top ", ntop),
                   filename = file.path(NFATc1_mut_vs_WT_results_motif_allPeaks_dir, paste0("motif_enrichment_signifTop", ntop,".pdf")))
# 
# pheatmap::pheatmap(devToPlot, 
#                    scale = "none", 
#                    cluster_cols = FALSE,
#                    annotation_col = annotCol_forHeatmap,
#                    main = paste0("Heatmap of motif enrichment - top ", ntop),
#                    filename = file.path(out_dir, paste0("motif_enrichment_signifTop", ntop,".png")))
#motif_dev_heatmap <- ComplexHeatmap::pheatmap(as.matrix(devToPlot), scale="none", cluster_cols = FALSE)

# replacing heatmap with publication heatmap ----
library("RColorBrewer")
annotCol_forHeatmap_colors <- list(genotype = c(  WT="#525252", `NFATc1_mut` = "#fc4e2a"),
                                   bio_rep = c(`ND_RJH` = "#d9d9d9", `ND_MK` = "#969696", `PID1185` = "#fd8d3c", `PID1180` = "#fd8d3c"))
#motifs_heatmap <-  
pheatmap::pheatmap(as.matrix(devToPlot),
                   color = colorRampPalette(rev(brewer.pal(7, "PiYG")))(100),
                   scale = "row",
                   #gaps_col = ph_col_annot%>% arrange(state)%$% table(state) %>% cumsum,
                   #gaps_row = ph_genes$clust %>% table() %>% cumsum,
                   cluster_cols = FALSE,
                   cluster_rows = TRUE,
                   show_colnames = FALSE,
                   show_rownames = TRUE,
                   annotation_col = annotCol_forHeatmap,
                   #annotation_row = ph_row_annot,
                   #annotation_legend = FALSE,
                   annotation_colors = annotCol_forHeatmap_colors,  
                   #labels_row = row_labels,
                   fontsize_row = 9,
                   fontsize = 9)

pheatmap::pheatmap(as.matrix(devToPlot),
                   color = colorRampPalette(rev(brewer.pal(7, "PiYG")))(100),
                   scale = "row",
                   #gaps_col = ph_col_annot%>% arrange(state)%$% table(state) %>% cumsum,
                   #gaps_row = ph_genes$clust %>% table() %>% cumsum,
                   cluster_cols = FALSE,
                   cluster_rows = TRUE,
                   show_colnames = FALSE,
                   show_rownames = TRUE,
                   annotation_col = annotCol_forHeatmap,
                   #annotation_row = ph_row_annot,
                   #annotation_legend = FALSE,
                   annotation_colors = annotCol_forHeatmap_colors,  
                   #labels_row = row_labels,
                   fontsize_row = 9,
                   fontsize = 9,
                   #cellheight = 9,
                   main = paste0("Heatmap of motif enrichment - top ", ntop),
                   filename = file.path(NFATc1_mut_vs_WT_results_motif_allPeaks_dir, paste0("motif_enrichment_signifTop", ntop,".pdf")))

pheatmap::pheatmap(as.matrix(devToPlot),
                   color = colorRampPalette(rev(brewer.pal(7, "PiYG")))(100),
                   scale = "row",
                   #gaps_col = ph_col_annot%>% arrange(state)%$% table(state) %>% cumsum,
                   #gaps_row = ph_genes$clust %>% table() %>% cumsum,
                   cluster_cols = FALSE,
                   cluster_rows = TRUE,
                   show_colnames = FALSE,
                   show_rownames = TRUE,
                   annotation_col = annotCol_forHeatmap,
                   #annotation_row = ph_row_annot,
                   #annotation_legend = FALSE,
                   annotation_colors = annotCol_forHeatmap_colors,  
                   #labels_row = row_labels,
                   fontsize_row = 9,
                   fontsize = 9,
                   #cellheight = 9,
                   main = paste0("Heatmap of motif enrichment - top ", ntop),
                   filename = file.path(NFATc1_mut_vs_WT_results_motif_allPeaks_dir, paste0("motif_enrichment_signifTop", ntop,".png")))

# pdf(file.path(out_dir, paste0("motif_enrichment_signifTop", ntop,".pdf")))
# grid.draw(motifs_heatmap)
# dev.off()

#TODO:
# do actuall stats test that there is significant different in variability between my group of interest
motif_results_list <- list(motif_variab_raw = chrom_access_variability_ord_results,
                           motif_variab_signif = chrom_access_variability_ord_results_directions,
                           motif_variab_signif_direction = chrom_access_variability_ord_results_directions_signifConsistPatients)
openxlsx::write.xlsx(x = motif_results_list, 
                     file = paste0(NFATc1_mut_vs_WT_results_motif_allPeaks_dir, "/motif_enrichment.xlsx"),
                     asTable = TRUE,
                     overwrite = TRUE) 


```
